---
title: "Appendix: Polynomial Structure and Extrapolation"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 7, dpi = 150)
library(dplyr)
library(ggplot2)
```

```{r load-data}
to  <- readRDS("../results_summary/thinned_oracle_s50.RDS") %>%
  filter(n_reps_used == 1)
res <- readRDS("../results_summary/equal_allocation_results.RDS")

oracle_full <- res$oracle %>%
  select(comp_no, config, nbasis, full_mse = mse_true)

# Per-sample gap
gap_df <- to %>%
  left_join(oracle_full, by = c("comp_no", "config", "nbasis")) %>%
  mutate(gap = thinned_mse - full_mse,
         t = (1 - epsilon) / epsilon) %>%
  filter(!is.na(gap))

# Seed-averaged gap
gap_avg <- gap_df %>%
  group_by(config, nbasis, epsilon, t) %>%
  summarise(mean_gap = mean(gap), .groups = "drop")

config_labels <- c("equal_30" = "n = 30", "equal_40" = "n = 40",
                   "equal_50" = "n = 50", "equal_75" = "n = 75",
                   "equal_100" = "n = 100", "equal_125" = "n = 125")

nb_show <- c(6, 18, 30, 42)
```

# Appendix content

Everything above the "Extra analysis" section below maps directly to the
paper appendix (Section `\ref{app:extrapolation}`).

## 1. Polynomial Structure of the Gap

Under the Fay--Herriot model with known $\beta$ and $\sigma^2$
(Proposition 3.1 conditions), the per-area thinning gap is
$$
\Delta_i(\varepsilon) = \frac{1 - \varepsilon}{\varepsilon}\, \gamma_i(\varepsilon)\, \gamma_i\, d_i.
$$
Reparametrize to $t = (1 - \varepsilon)/\varepsilon$, the odds of
withheld-to-retained data, so that $t = 0$ corresponds to the
full-data limit $\varepsilon = 1$. Defining
$\alpha_i = 1 - \gamma_i = d_i / (\sigma^2 + d_i)$ and substituting
$\varepsilon = 1/(1+t)$, the gap simplifies to
$$
\Delta_i(\varepsilon) = \gamma_i^2\, d_i \cdot \frac{t}{1 + \alpha_i\, t}.
$$
This is a ratio of linear polynomials in $t$: it vanishes at $t = 0$,
is monotonically increasing and concave, and is well-approximated by
a low-order polynomial for moderate $t$. The $g_{2i}(\varepsilon)$
correction from Proposition 3.2 (estimated $\beta$) is $O(p/m)$ and
itself rational in $t$, so the polynomial structure is preserved
under estimated parameters.


## 2. Empirical Validation

We use the **oracle thinned MSE** (Theorem 1 estimand, evaluated
at the true $\theta_i$ rather than the test-set proxy).

### Average gap vs $t$

The oracle thinning gap averaged over $S = 50$ simulated samples,
plotted against $t = (1 - \varepsilon)/\varepsilon$ for four
representative basis counts: $p = 6$ (simple), $p = 18$ and $p = 30$
(moderate), $p = 42$ (complex). We restrict to $\varepsilon \geq 0.3$
($t \leq 2.33$), the range spanned by the extrapolation grids; the
saturating large-$t$ tail ($\varepsilon = 0.1, 0.2$) is never used and
is omitted. Over this range the gap is approximately linear in $t$,
heading to $0$ at $t = 0$ ($\varepsilon = 1$), consistent with the
leading $\Delta \approx c\,t$ term of the polynomial structure from
Section 1. Each point is one (design, $p$, $\varepsilon$) cell; the
line is the OLS fit.

```{r plot-avg-gap, fig.height=8}
gap_line <- gap_avg %>%
  filter(nbasis %in% nb_show, epsilon >= 0.3) %>%   # t <= 2.33, the extrapolation-grid range
  mutate(config_lab = config_labels[config],
         nbasis_lab = factor(paste0("p = ", nbasis),
                             levels = paste0("p = ", sort(nb_show))))

ggplot(gap_line, aes(x = t, y = mean_gap, color = nbasis_lab)) +
  geom_point(size = 2.5) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, linewidth = 0.8) +
  facet_wrap(~ config_lab, scales = "free_y", nrow = 2) +
  scale_color_brewer(palette = "Set1", name = "Model") +
  labs(x = expression(t == (1 - epsilon) / epsilon),
       y = "Oracle thinning gap") +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank(),
        legend.position = "bottom",
        strip.text = element_text(face = "bold"))

ggsave("plot_appendix_extrapolation.png",
       width = 10, height = 7, dpi = 300)
```

On the original $t$ vs gap axes the averaged gap is approximately
linear across all designs and model complexities, heading to the
origin, consistent with the $\Delta \approx c\,t$ leading term of the
polynomial structure from Section 1. More complex models ($p = 42$)
have larger gaps, as expected from the theory.

### Pooled extrapolation estimator

The polynomial structure suggests estimating
$\text{MSE}_\text{full}$ by evaluating
$\widehat{\text{MSE}}_\varepsilon$ at a grid
$\varepsilon_1, \ldots, \varepsilon_K$, fitting
$$
\widehat{\text{MSE}}_{\varepsilon_k} = \alpha_0 + \alpha_1\, t_k + \alpha_2\, t_k^2 + \eta_k,
$$
and taking the intercept $\hat\alpha_0$ as the extrapolation to
$t = 0$ ($\varepsilon = 1$). Under oracle conditions (known
$\theta_i$, no test-set noise), this estimator agrees with the oracle
$p^*$ in all 6 equal-allocation designs, confirming the polynomial
functional form.

## 3. Feasible Extrapolation Fails

The feasible $\widehat{\text{MSE}}_\varepsilon$ (Theorem 1) replaces
$\theta_i$ with the test-set proxy
$z_i^{(2)} / (1 - \varepsilon)$, introducing noise of variance
$d_i / (1 - \varepsilon)$ per area. For each sample, the pooled
estimator fits the regression using $R = 5$ thinning
replicates at each grid point and selects the basis count with the
smallest intercept $\hat\alpha_0$. We test both quadratic and linear
fits over two grids:
$\varepsilon \in \{0.3, 0.4, 0.5, 0.6, 0.7\}$ ($K = 5$) and
$\varepsilon \in \{0.5, 0.6, 0.7\}$ ($K = 3$).
The baseline is single-$\varepsilon$ selection (pick the basis count
minimizing $\widehat{\text{MSE}}_\varepsilon$ at a fixed $\varepsilon$).
The evaluation metric is RMSE of the selected basis count relative to
the per-sample oracle $p^*$.

```{r feasible-table}
N_REPS <- 5
LOSS   <- "MSE"

dt_all <- res$dt_1fold %>%
  filter(loss_function == LOSS, n_reps_used == N_REPS) %>%
  mutate(epsilon = round(epsilon, 1), u = (1 - epsilon) / epsilon) %>%
  select(config, comp_no, nbasis, epsilon, u, mse_hat = metric)

# Average oracle p* (matches main paper definition)
oracle_pstar <- res$oracle %>%
  group_by(config, nbasis) %>%
  summarise(avg_mse = mean(mse_true), .groups = "drop") %>%
  group_by(config) %>%
  slice_min(avg_mse, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(config, oracle_p = nbasis)

rmse_score <- function(picks) {
  picks %>%
    left_join(oracle_pstar, by = "config") %>%
    group_by(config) %>%
    summarise(rmse_p = sqrt(mean((nbasis - oracle_p)^2)), .groups = "drop")
}

# Pooled estimators (OLS)
run_pooled <- function(grid, degree) {
  sub <- dt_all %>% filter(epsilon %in% grid)
  intercepts <- sub %>%
    group_by(config, comp_no, nbasis) %>%
    summarise(
      alpha = if (degree == "quadratic") {
        coef(lm(mse_hat ~ u + I(u^2)))[1]
      } else {
        coef(lm(mse_hat ~ u))[1]
      },
      .groups = "drop")
  intercepts %>%
    group_by(config, comp_no) %>%
    slice_min(alpha, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    rmse_score()
}

# Single-epsilon baselines
run_single <- function(e) {
  dt_all %>%
    filter(abs(epsilon - e) < 0.01) %>%
    group_by(config, comp_no) %>%
    slice_min(mse_hat, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    rmse_score()
}

methods <- list(
  list(label = "Quadratic [0.3, 0.7]", fn = function() run_pooled(seq(0.3, 0.7, 0.1), "quadratic")),
  list(label = "Linear [0.3, 0.7]",    fn = function() run_pooled(seq(0.3, 0.7, 0.1), "linear")),
  list(label = "Linear [0.5, 0.7]",    fn = function() run_pooled(seq(0.5, 0.7, 0.1), "linear"))
)

baselines <- list(
  list(label = "eps = 0.5", fn = function() run_single(0.5)),
  list(label = "eps = 0.6", fn = function() run_single(0.6)),
  list(label = "eps = 0.7", fn = function() run_single(0.7))
)

configs <- c("equal_30", "equal_40", "equal_50",
             "equal_75", "equal_100", "equal_125")

build_row <- function(m) {
  r <- m$fn()
  vals <- sapply(configs, function(cf) {
    v <- r$rmse_p[r$config == cf]
    if (length(v) == 0) NA else round(v, 1)
  })
  c(Method = m$label, setNames(vals, config_labels[configs]))
}

tbl <- bind_rows(
  lapply(methods, build_row) %>% bind_rows(),
  tibble(Method = "---", `n = 30` = "", `n = 40` = "", `n = 50` = "",
         `n = 75` = "", `n = 100` = "", `n = 125` = ""),
  lapply(baselines, build_row) %>% bind_rows()
)

knitr::kable(tbl,
             caption = "RMSE of selected nbasis vs per-sample oracle p* (R = 5, DT-MSE)",
             align = c("l", rep("r", 6)))
```

The pooled estimator underperforms every single-$\varepsilon$ baseline
across all 6 designs. The degradation is largest at high sample sizes
($n = 100$, $n = 125$) where the true model spread is smallest and
the extrapolation noise is most damaging relative to the signal.

Linear and quadratic pooled fits perform similarly, confirming that the
issue is extrapolation distance to $t = 0$, not polynomial degree.


---

# Extra analysis

Additional plots and tables not included in the paper appendix.
Retained for reproducibility and deeper investigation.

## Per-sample gap vs $t$

The same oracle thinning gap, but now showing per-sample linear fits.
Each line is the OLS fit for a single simulated sample ($S = 50$
lines per panel). This shows whether the linear relationship holds
sample-by-sample, not just on average.

```{r plot-persample-gap, fig.height=8}
gap_spaghetti <- gap_df %>%
  filter(nbasis %in% nb_show) %>%
  mutate(config_lab = config_labels[config],
         nbasis_lab = paste0("p = ", nbasis))

ggplot(gap_spaghetti, aes(x = t, y = gap,
                           group = interaction(comp_no, nbasis_lab),
                           color = nbasis_lab)) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE,
              alpha = 0.3, linewidth = 0.3) +
  facet_wrap(~ config_lab, scales = "free_y", nrow = 2) +
  scale_color_brewer(palette = "Set1", name = "Model") +
  labs(x = expression(t == (1 - epsilon) / epsilon),
       y = "Oracle thinning gap (per-sample linear fits)") +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank(),
        legend.position = "bottom",
        strip.text = element_text(face = "bold"))
```

The linear relationship holds per-sample: individual fitted lines are
approximately parallel, confirming the polynomial structure is not an
artifact of averaging. The vertical spread across samples reflects
the across-sample variance --- the variability due to which particular
sample $y$ was drawn from the population.


## Oracle pooled estimator: detailed tables

### Seed-averaged oracle agreement

```{r oracle-avg}
pool_dat <- gap_avg %>%
  left_join(oracle_full %>%
              group_by(config, nbasis) %>%
              summarise(mean_full = mean(full_mse), .groups = "drop"),
            by = c("config", "nbasis")) %>%
  mutate(mean_thin = mean_gap + mean_full)

oracle_pick_avg <- pool_dat %>%
  select(config, nbasis, mean_full) %>%
  distinct() %>%
  group_by(config) %>%
  slice_min(mean_full, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(config, oracle_p = nbasis)

pooled_pick_avg <- pool_dat %>%
  group_by(config, nbasis) %>%
  summarise(alpha_hat = coef(lm(mean_thin ~ t + I(t^2)))[1],
            .groups = "drop") %>%
  group_by(config) %>%
  slice_min(alpha_hat, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(config, pooled_p = nbasis)

comp_avg <- oracle_pick_avg %>%
  left_join(pooled_pick_avg, by = "config") %>%
  mutate(match = oracle_p == pooled_p,
         config_lab = config_labels[config])

knitr::kable(comp_avg %>% select(Design = config_lab,
                                  `Oracle p*` = oracle_p,
                                  `Pooled p*` = pooled_p,
                                  Match = match),
             caption = "Seed-averaged oracle: pooled estimator vs oracle p*")
```

### Per-sample oracle RMSE

```{r oracle-persample}
oracle_thin_persample <- gap_df %>%
  mutate(thinned_mse_val = gap + full_mse) %>%
  select(config, comp_no, nbasis, epsilon, t, thinned_mse = thinned_mse_val)

oracle_pstar_persample <- oracle_full %>%
  group_by(config, comp_no) %>%
  slice_min(full_mse, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(config, comp_no, oracle_p = nbasis)

pooled_persample <- oracle_thin_persample %>%
  group_by(config, comp_no, nbasis) %>%
  summarise(alpha_hat = coef(lm(thinned_mse ~ t + I(t^2)))[1],
            .groups = "drop") %>%
  group_by(config, comp_no) %>%
  slice_min(alpha_hat, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(config, comp_no, pooled_p = nbasis)

single_oracle <- function(eps_val) {
  oracle_thin_persample %>%
    filter(abs(epsilon - eps_val) < 0.01) %>%
    group_by(config, comp_no) %>%
    slice_min(thinned_mse, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    select(config, comp_no, sel_p = nbasis) %>%
    left_join(oracle_pstar_persample, by = c("config", "comp_no")) %>%
    group_by(config) %>%
    summarise(rmse = sqrt(mean((sel_p - oracle_p)^2)), .groups = "drop")
}

pooled_score <- pooled_persample %>%
  left_join(oracle_pstar_persample, by = c("config", "comp_no")) %>%
  group_by(config) %>%
  summarise(rmse = sqrt(mean((pooled_p - oracle_p)^2)), .groups = "drop") %>%
  mutate(method = "Pooled quadratic")

single_scores <- bind_rows(
  single_oracle(0.5) %>% mutate(method = "eps = 0.5"),
  single_oracle(0.6) %>% mutate(method = "eps = 0.6"),
  single_oracle(0.7) %>% mutate(method = "eps = 0.7")
)

oracle_tbl <- bind_rows(pooled_score, single_scores) %>%
  tidyr::pivot_wider(names_from = config, values_from = rmse) %>%
  select(Method = method, all_of(names(config_labels)))

names(oracle_tbl)[-1] <- config_labels[names(oracle_tbl)[-1]]
oracle_tbl[-1] <- round(oracle_tbl[-1], 1)

knitr::kable(oracle_tbl,
             caption = "Per-sample oracle: RMSE of selected p vs per-sample oracle p*",
             align = c("l", rep("r", 6)))
```

Under oracle conditions (known $\theta_i$), the per-sample pooled
estimator should still benefit from the correct polynomial structure.
The comparison to single-$\varepsilon$ baselines shows how much the
extrapolation helps (or doesn't) even before test-set noise enters.
