---
title: "Section 6.3: Methods Comparison on Proportional Allocation Designs"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6)
library(tidyverse)
library(knitr)
library(xtable)
```

# Overview

Comparison of model selection methods on **proportional allocation (PA) designs**:

- **Designs:** 0.75%, 1.25%, 1.75% sampling rates
- **Methods:** DT-MSE, DT-Likelihood (ε = 0.6, R=5), DIC, WAIC, ESIM
- **Models:** p ∈ {3, 6, 9, ..., 60} spatial basis functions (20 candidate models)
- **Samples:** S = 50 per design

---

# Data Loading

```{r load_data}
results <- readRDS("../results_summary/methodcomp_results.RDS")

oracle_all <- results$oracle  # comp_no, model, nbasis, mse_true, WAIC, DIC, config
dt_all     <- results$dt      # comp_no, method, model, metric, loss_function, nbasis, epsilon, n_reps_used, config
esim_all   <- results$esim    # comp_no, method, model, metric, nbasis, config

# Oracle selections (argmin mse_true per comparison)
oracle_selections <- oracle_all %>%
  filter(!is.na(mse_true)) %>%
  group_by(config, comp_no) %>%
  slice_min(mse_true, n = 1, with_ties = FALSE) %>%
  select(config, comp_no, oracle_nbasis = nbasis) %>%
  ungroup()

# DIC / WAIC selections
ic_selections <- oracle_all %>%
  pivot_longer(cols = c(DIC, WAIC), names_to = "method_label", values_to = "ic_val") %>%
  filter(!is.na(ic_val)) %>%
  group_by(config, comp_no, method_label) %>%
  slice_min(ic_val, n = 1, with_ties = FALSE) %>%
  select(config, comp_no, method_label, selected_nbasis = nbasis) %>%
  mutate(method_type = method_label, epsilon = NA_real_) %>%
  ungroup()

# DT selections (R=5 only, epsilon = 0.6)
dt_selections <- dt_all %>%
  filter(n_reps_used == 5,
         abs(epsilon - 0.6) < 0.001) %>%
  mutate(
    method_type = case_when(
      str_detect(method, "MSE") ~ "DT-MSE",
      str_detect(method, "plugin_NLL") ~ "DT-NLL"
    ),
    method_label = sprintf("%s \u03b5=%.1f", method_type, epsilon)
  ) %>%
  group_by(config, comp_no, method_label, method_type, epsilon) %>%
  slice_min(metric, n = 1, with_ties = FALSE) %>%
  select(config, comp_no, method_label, method_type, epsilon, selected_nbasis = nbasis) %>%
  ungroup()

# ESIM selections (standard validation only)
esim_selections <- esim_all %>%
  filter(method == "ESIM_standard", metric_type == "MSE", !is.na(nbasis)) %>%
  group_by(config, comp_no) %>%
  slice_min(metric, n = 1, with_ties = FALSE) %>%
  select(config, comp_no, selected_nbasis = nbasis) %>%
  mutate(method_label = "ESIM", method_type = "ESIM", epsilon = NA_real_) %>%
  ungroup()

# Combine and compute deviations
deviations_all <- bind_rows(dt_selections, ic_selections, esim_selections) %>%
  left_join(oracle_selections, by = c("config", "comp_no")) %>%
  mutate(
    deviation = selected_nbasis - oracle_nbasis,
    abs_deviation = abs(deviation)
  )

cat("Loaded data:\n")
cat(sprintf("  - Oracle selections: %d samples × %d designs = %d\n",
            n_distinct(oracle_selections$comp_no),
            n_distinct(oracle_selections$config),
            nrow(oracle_selections)))
cat(sprintf("  - Deviations: %d rows\n", nrow(deviations_all)))
cat(sprintf("  - Methods: %s\n", paste(sort(unique(deviations_all$method_label)), collapse = ", ")))
```

---

# Oracle Basis Statistics

```{r oracle_stats}
# Oracle basis statistics by design
oracle_stats <- oracle_selections %>%
  group_by(config) %>%
  summarise(
    n_samples = n(),
    mean_oracle = mean(oracle_nbasis),
    sd_oracle = sd(oracle_nbasis),
    min_oracle = min(oracle_nbasis),
    max_oracle = max(oracle_nbasis),
    range_oracle = max(oracle_nbasis) - min(oracle_nbasis)
  ) %>%
  mutate(
    config_label = case_when(
      config == "prop_0.75pct" ~ "0.75% PA",
      config == "prop_1p25pct" ~ "1.25% PA",
      config == "prop_1p75pct" ~ "1.75% PA"
    )
  ) %>%
  select(config_label, mean_oracle, sd_oracle, min_oracle, max_oracle, range_oracle)

oracle_stats %>%
  kable(digits = 1,
        col.names = c("Design", "Mean", "SD", "Min", "Max", "Range"),
        caption = "Oracle basis function counts by design (across 50 samples)")

# Print oracle basis statistics
cat("\n\nOracle basis function statistics (S=50):\n")
print(oracle_stats)
```

---

# Output 1: MAE and Bias Comparison Table

```{r comparison_table}
# Filter to methods for table
table_methods <- c(
  "DT-MSE ε=0.6",
  "DT-NLL ε=0.6",
  "DIC",
  "WAIC",
  "ESIM"
)

config_order <- c("prop_0.75pct", "prop_1p25pct", "prop_1p75pct")
config_pct   <- c("0.75", "1.25", "1.75")

# Compute MAE and bias by method and design
comp_stats <- deviations_all %>%
  filter(method_label %in% table_methods) %>%
  group_by(method_label, config) %>%
  summarise(MAE = mean(abs_deviation), bias = mean(deviation), .groups = "drop")

# Fixed order: DT-first, then info criteria, then ESIM
method_order <- c("DT-MSE \u03b5=0.6", "DT-NLL \u03b5=0.6", "DIC", "WAIC", "ESIM")

# Display HTML version
comp_stats %>%
  mutate(cell = sprintf("MAE=%.2f, bias=%+.1f", MAE, bias)) %>%
  select(method_label, config, cell) %>%
  pivot_wider(names_from = config, values_from = cell) %>%
  kable(caption = "MAE and bias from oracle basis count (S=50)")

# Build LaTeX table manually (xtable can't do multi-column headers with cmidrule)
format_bias <- function(b) {
  s <- sprintf("%.1f", abs(b))
  if (b >= 0) paste0("$+$", s) else paste0("$-$", s)
}

latex_lines <- c(
  "\\begin{table}[ht]",
  "\\centering",
  "\\begin{tabular}{l rrr rrr}",
  "  \\toprule",
  "  & \\multicolumn{3}{c}{MAE} & \\multicolumn{3}{c}{Bias} \\\\",
  "  \\cmidrule(lr){2-4} \\cmidrule(lr){5-7}",
  sprintf("  Method & %s \\\\",
          paste(sprintf("%s\\%%", config_pct), collapse = " & ") %>%
          { paste0(., " & ", paste(sprintf("%s\\%%", config_pct), collapse = " & ")) })
)
latex_lines <- c(latex_lines, "  \\midrule")

for (m in method_order) {
  m_latex <- str_replace(m, "ε", "$\\\\epsilon$")
  row_data <- comp_stats %>% filter(method_label == m)

  mae_cells <- sapply(config_order, function(cfg) {
    val <- row_data %>% filter(config == cfg) %>% pull(MAE)
    sprintf("%.2f", val)
  })

  bias_cells <- sapply(config_order, function(cfg) {
    val <- row_data %>% filter(config == cfg) %>% pull(bias)
    format_bias(val)
  })

  latex_lines <- c(latex_lines,
    sprintf("  %s & %s & %s \\\\", m_latex,
            paste(mae_cells, collapse = " & "),
            paste(bias_cells, collapse = " & ")))
}

latex_lines <- c(latex_lines,
  "  \\bottomrule",
  "\\end{tabular}",
  sprintf("\\caption{MAE and mean bias from oracle basis count by method and design across $S=50$ simulation replicates. Columns indicate proportional-to-population allocation (PA) designs with overall sampling rates of %s. Negative bias indicates under-selection; positive indicates over-selection.}",
          paste(sprintf("%s\\%%", config_pct), collapse = ", ")),
  "\\label{tab:comparison}",
  "\\end{table}")

writeLines(latex_lines, "table_comparison.tex")
cat("\nGenerated LaTeX:\n")
cat(paste(latex_lines, collapse = "\n"))
```

---

# Output 2: Boxplot of Deviations

```{r deviation_boxplot, fig.width=10, fig.height=4}
# Filter to keep only ε=0.6 for DT methods
deviations_plot <- deviations_all %>%
  filter(is.na(epsilon) | abs(epsilon - 0.6) < 0.001) %>%
  mutate(
    config_label = case_when(
      config == "prop_0.75pct" ~ "0.75% PA",
      config == "prop_1p25pct" ~ "1.25% PA",
      config == "prop_1p75pct" ~ "1.75% PA"
    ),
    # Set factor order for x-axis
    method_label = factor(method_label, levels = c(
      "DT-MSE \u03b5=0.6",
      "DT-NLL \u03b5=0.6",
      "DIC",
      "WAIC",
      "ESIM"
    )),
    # Categorize as DT vs non-DT
    method_category = ifelse(method_type %in% c("DT-MSE", "DT-NLL"), "DT", "Non-DT")
  )

# Create boxplot
p <- ggplot(deviations_plot, aes(x = method_label, y = deviation, fill = method_category)) +
  geom_boxplot(alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  facet_wrap(~config_label, ncol = 3) +
  labs(
    x = "Method",
    y = "Deviation from oracle basis"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 11),
    axis.text.y = element_text(size = 11),
    axis.title = element_text(size = 12),
    strip.text = element_text(size = 12),
    panel.grid.minor = element_blank(),
    legend.position = "none"
  ) +
  scale_fill_manual(values = c("DT" = "lightblue", "Non-DT" = "lightgray"))

# Save plot
ggsave("plot5_comparison.png", p, width = 10, height = 4, dpi = 300)
```

---

# Output 3: Responsiveness to Oracle Signal (1.75% PA)

```{r responsiveness_table}
# Compute oracle margins per sample (gap between best and 2nd-best model MSE)
oracle_margins <- oracle_all %>%
  filter(!is.na(mse_true)) %>%
  group_by(config, comp_no) %>%
  arrange(mse_true) %>%
  summarise(
    margin_1v2 = mse_true[2] - mse_true[1],
    .groups = "drop"
  )

# Focus on 1.75% PA
margins_175 <- oracle_margins %>% filter(config == "prop_1p75pct")
median_margin <- median(margins_175$margin_1v2)

# Identify clear-signal samples (above median margin)
clear_signal_comps <- margins_175 %>%
  filter(margin_1v2 >= median_margin) %>%
  pull(comp_no)

n_all   <- n_distinct(margins_175$comp_no)
n_clear <- length(clear_signal_comps)

# Deviations for 1.75% only
dev_175 <- deviations_all %>%
  filter(config == "prop_1p75pct", method_label %in% table_methods)

# MAE for all samples
mae_all <- dev_175 %>%
  group_by(method_label) %>%
  summarise(MAE_all = mean(abs_deviation), .groups = "drop")

# MAE for clear-signal samples
mae_clear <- dev_175 %>%
  filter(comp_no %in% clear_signal_comps) %>%
  group_by(method_label) %>%
  summarise(MAE_clear = mean(abs_deviation), .groups = "drop")

# Combine and compute change
responsiveness <- mae_all %>%
  left_join(mae_clear, by = "method_label") %>%
  mutate(pct_change = (MAE_clear - MAE_all) / MAE_all * 100) %>%
  arrange(MAE_all)

# Display HTML version
responsiveness %>%
  mutate(change_str = sprintf("%+.0f%%", round(pct_change))) %>%
  select(method_label, MAE_all, MAE_clear, change_str) %>%
  kable(digits = 2, col.names = c("Method", "All (S=50)", "Clear Signal (S=25)", "Change"),
        caption = sprintf("Responsiveness to oracle signal at 1.75%% PA (median margin = %.2e)", median_margin))

# Replaced by scatter plot (Output 4) -- table_responsiveness.tex no longer needed
# format_change <- function(pct) {
#   if (pct >= 0) sprintf("$+$%.0f\\%%", round(abs(pct)))
#   else sprintf("$-$%.0f\\%%", round(abs(pct)))
# }
#
# resp_lines <- c(
#   "\\begin{table}[ht]",
#   "\\centering",
#   "\\begin{tabular}{lrrr}",
#   "  \\toprule",
#   sprintf("  Method & All ($S$=%d) & Clear Signal ($S$=%d) & Change \\\\", n_all, n_clear),
#   "  \\midrule"
# )
#
# for (i in 1:nrow(responsiveness)) {
#   r <- responsiveness[i, ]
#   m_latex <- str_replace(r$method_label, "ε", "$\\\\epsilon$")
#   resp_lines <- c(resp_lines,
#     sprintf("  %s & %.2f & %.2f & %s \\\\",
#             m_latex, r$MAE_all, r$MAE_clear, format_change(r$pct_change)))
# }
#
# resp_lines <- c(resp_lines,
#   "  \\bottomrule",
#   "\\end{tabular}",
#   sprintf("\\caption{...}", n_clear, n_all),
#   "\\label{tab:responsiveness}",
#   "\\end{table}")
#
# writeLines(resp_lines, "table_responsiveness.tex")
```

---

# Output 4: Oracle Margin Scatter Plot (1.75% PA)

```{r oracle_margin_scatter, fig.width=12, fig.height=4}
# Build per-sample scatter data: x = oracle margin, y = abs deviation
scatter_dat <- deviations_all %>%
  filter(
    config == "prop_1p75pct",
    method_label %in% table_methods,
    is.na(epsilon) | abs(epsilon - 0.6) < 0.001
  ) %>%
  left_join(
    oracle_margins %>% filter(config == "prop_1p75pct"),
    by = c("config", "comp_no")
  ) %>%
  mutate(
    method_clean = case_when(
      method_label == "DT-MSE \u03b5=0.6" ~ "DT-MSE",
      method_label == "DT-NLL \u03b5=0.6" ~ "DT-NLL",
      TRUE ~ method_label
    ),
    method_clean = factor(method_clean, levels = c("DT-MSE", "DT-NLL", "DIC", "WAIC", "ESIM")),
    margin_scaled = margin_1v2 * 1e5
  )

p_scatter <- ggplot(scatter_dat, aes(x = margin_scaled, y = abs_deviation)) +
  geom_point(alpha = 0.45, size = 1.8, color = "gray35") +
  geom_smooth(method = "loess", se = FALSE,
              color = "steelblue4", linewidth = 0.9) +
  facet_wrap(~method_clean, nrow = 1) +
  labs(
    x = expression("Oracle margin (" %*% 10^{-5} * "), MSE gap: best vs. 2nd-best model"),
    y = "Absolute deviation from oracle"
  ) +
  scale_y_continuous(breaks = seq(0, 60, by = 9)) +
  theme_bw() +
  theme(
    strip.text = element_text(size = 12, face = "bold"),
    strip.background = element_rect(fill = "gray95"),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 11),
    panel.grid.minor = element_blank()
  )

print(p_scatter)
ggsave("plot6_oracle_margin.png", p_scatter, width = 12, height = 4, dpi = 300)
```

---

# Session Info

```{r session_info}
sessionInfo()
```
