---
title: "Section 6.3: Methods Comparison on Proportional Allocation Designs"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6)
library(tidyverse)
library(knitr)
```

# Overview

Comparison of model selection methods on **proportional allocation (PA) designs**:

- **Designs:** 0.75%, 1.25%, 1.75% sampling rates
- **Methods:** DT-MSE, DT-NLL (ε = 0.6, R=5), DIC, WAIC, ESIM
- **Models:** p ∈ {3, 6, 9, ..., 60} spatial basis functions (20 candidate models)
- **Samples:** S = 50 per design
- **Oracle:** Average oracle (argmin_p mean_s MSE(p)) = p = 15 for all PA designs
- **Metric:** RMSE from average oracle

---

# Data Loading

```{r load_data}
results <- readRDS("../results_summary/methodcomp_results.RDS")

oracle_all <- results$oracle
dt_all     <- results$dt
esim_all   <- results$esim

CONFIGS <- c("prop_0.75pct", "prop_1p25pct", "prop_1p75pct")
config_labels <- c(prop_0.75pct = "0.75% PA", prop_1p25pct = "1.25% PA", prop_1p75pct = "1.75% PA")

# Average oracle
avg_oracle <- oracle_all %>%
  filter(config %in% CONFIGS, !is.na(mse_true)) %>%
  group_by(config, nbasis) %>%
  summarise(avg_mse = mean(mse_true), .groups = "drop") %>%
  group_by(config) %>%
  slice_min(avg_mse, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(config, avg_oracle = nbasis)

cat("Average oracle by design:\n")
print(avg_oracle)
```

```{r selections}
EPS <- results$metadata$epsilon  # 0.6

# DT selections (MSE and NLL)
dt_mse_sel <- dt_all %>%
  filter(config %in% CONFIGS, loss_function == "MSE",
         abs(epsilon - EPS) < 0.001, n_reps_used == 5) %>%
  group_by(config, comp_no) %>%
  slice_min(metric, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(method = "DT-MSE") %>%
  select(config, comp_no, method, selected = nbasis)

dt_nll_sel <- dt_all %>%
  filter(config %in% CONFIGS, loss_function == "plugin_NLL",
         abs(epsilon - EPS) < 0.001, n_reps_used == 5) %>%
  group_by(config, comp_no) %>%
  slice_min(metric, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(method = "DT-NLL") %>%
  select(config, comp_no, method, selected = nbasis)

# DIC / WAIC selections
dic_sel <- oracle_all %>%
  filter(config %in% CONFIGS) %>%
  group_by(config, comp_no) %>%
  slice_min(DIC, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(method = "DIC") %>%
  select(config, comp_no, method, selected = nbasis)

waic_sel <- oracle_all %>%
  filter(config %in% CONFIGS) %>%
  group_by(config, comp_no) %>%
  slice_min(WAIC, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(method = "WAIC") %>%
  select(config, comp_no, method, selected = nbasis)

# ESIM selections
esim_sel <- esim_all %>%
  filter(config %in% CONFIGS, method == "ESIM_standard",
         metric_type == "MSE", !is.na(nbasis)) %>%
  group_by(config, comp_no) %>%
  slice_min(metric, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(method = "ESIM") %>%
  select(config, comp_no, method, selected = nbasis)

# Combine and compute deviations from average oracle
all_sel <- bind_rows(dt_mse_sel, dt_nll_sel, dic_sel, waic_sel, esim_sel) %>%
  left_join(avg_oracle, by = "config") %>%
  mutate(dev = selected - avg_oracle)

cat(sprintf("Total selections: %d (%d methods × %d designs × %d samples)\n",
            nrow(all_sel), n_distinct(all_sel$method),
            n_distinct(all_sel$config), 50))
```

---

# Output 1: RMSE and Bias Table

```{r comparison_table}
method_order <- c("DT-MSE", "DT-NLL", "DIC", "WAIC", "ESIM")

# RMSE by method and design
rmse_tab <- all_sel %>%
  group_by(method, config) %>%
  summarise(RMSE = sqrt(mean(dev^2)), .groups = "drop") %>%
  mutate(config = config_labels[config]) %>%
  pivot_wider(names_from = config, values_from = RMSE)

# Overall RMSE
overall_rmse <- all_sel %>%
  group_by(method) %>%
  summarise(Overall = sqrt(mean(dev^2)), .groups = "drop")

cat("=== RMSE (average oracle) ===\n")
rmse_tab %>%
  left_join(overall_rmse, by = "method") %>%
  mutate(method = factor(method, levels = method_order)) %>%
  arrange(method) %>%
  kable(digits = 2)

# Bias by method and design
bias_tab <- all_sel %>%
  group_by(method, config) %>%
  summarise(Bias = mean(dev), .groups = "drop") %>%
  mutate(config = config_labels[config]) %>%
  pivot_wider(names_from = config, values_from = Bias)

overall_bias <- all_sel %>%
  group_by(method) %>%
  summarise(Overall = mean(dev), .groups = "drop")

cat("\n=== Mean Bias (average oracle) ===\n")
bias_tab %>%
  left_join(overall_bias, by = "method") %>%
  mutate(method = factor(method, levels = method_order)) %>%
  arrange(method) %>%
  kable(digits = 2)
```

```{r latex_table}
# Build LaTeX table: RMSE only
config_order <- c("prop_0.75pct", "prop_1p25pct", "prop_1p75pct")
config_pct <- c("0.75", "1.25", "1.75")

# Compute stats
comp_stats <- all_sel %>%
  group_by(method, config) %>%
  summarise(RMSE = sqrt(mean(dev^2)), .groups = "drop")

overall_rmse_stats <- all_sel %>%
  group_by(method) %>%
  summarise(RMSE_overall = sqrt(mean(dev^2)), .groups = "drop")

latex_lines <- c(
  "\\begin{table}[ht]",
  "\\centering",
  "\\begin{tabular}{l rrrr}",
  "  \\toprule",
  sprintf("  Method & %s & Overall \\\\",
          paste(sprintf("%s\\%%", config_pct), collapse = " & "))
)
latex_lines <- c(latex_lines, "  \\midrule")

dt_label <- sprintf("$\\\\epsilon$=%.1f", EPS)
method_latex_names <- c(
  "DT-MSE" = sprintf("DT-MSE %s", dt_label),
  "DT-NLL" = sprintf("DT-NLL %s", dt_label),
  "DIC" = "DIC", "WAIC" = "WAIC", "ESIM" = "ESIM"
)

for (m in method_order) {
  row_data <- comp_stats %>% filter(method == m)
  rmse_overall <- overall_rmse_stats %>% filter(method == m) %>% pull(RMSE_overall)
  rmse_cells <- sapply(config_order, function(cfg)
    sprintf("%.2f", row_data %>% filter(config == cfg) %>% pull(RMSE)))
  latex_lines <- c(latex_lines,
    sprintf("  %s & %s & %.2f \\\\", method_latex_names[m],
            paste(rmse_cells, collapse = " & "),
            rmse_overall))
}

latex_lines <- c(latex_lines,
  "  \\bottomrule",
  "\\end{tabular}",
  sprintf("\\caption{RMSE of the selected basis count from the average oracle ($p^* = 15$) by method and design across $S=50$ simulation replicates. Columns indicate proportional allocation designs with overall sampling rates of 0.75\\%%, 1.25\\%%, and 1.75\\%%. Overall RMSE pools deviations across all three designs.}"),
  "\\label{tab:comparison}",
  "\\end{table}")

writeLines(latex_lines, "table_comparison.tex")
cat("\nGenerated LaTeX:\n")
cat(paste(latex_lines, collapse = "\n"))
```

---

# Output 2: Boxplot of Selected Basis Functions

```{r boxplot, fig.width=10, fig.height=5}
plot_df <- all_sel %>%
  mutate(
    design = factor(config_labels[config], levels = config_labels),
    method = factor(method, levels = method_order)
  )

dt_methods <- c("DT-MSE", "DT-NLL")

p <- ggplot(plot_df, aes(x = method, y = selected)) +
  geom_hline(yintercept = 15, linetype = "dashed", color = "red", linewidth = 0.5) +
  geom_boxplot(aes(fill = method %in% dt_methods), outlier.size = 0.8, width = 0.6) +
  scale_fill_manual(values = c("TRUE" = "lightblue", "FALSE" = "gray90"), guide = "none") +
  facet_wrap(~ design, nrow = 1) +
  labs(x = NULL, y = "Selected number of basis functions",
       ) +
  theme_minimal() +
  theme(strip.text = element_text(size = 11, face = "bold"),
        axis.text.x = element_text(angle = 30, hjust = 1, size = 9))

print(p)
ggsave("plot5_comparison.png", p, width = 10, height = 5, dpi = 300)
```

---

# Output 2b: Histogram of Selected Basis Functions

```{r histogram, fig.width=10, fig.height=10, eval=FALSE}
method_order_rev <- rev(method_order)  # ESIM on top

p_hist <- ggplot(plot_df %>% mutate(method = factor(method, levels = method_order_rev)),
                 aes(x = selected, fill = method %in% dt_methods)) +
  geom_vline(xintercept = 15, linetype = "dashed", color = "red", linewidth = 0.5) +
  geom_bar() +
  scale_fill_manual(values = c("TRUE" = "#5B9BD5", "FALSE" = "gray50"), guide = "none") +
  scale_x_continuous(breaks = c(0, 15, 30, 45, 60)) +
  facet_grid(method ~ design) +
  labs(x = "Selected number of basis functions", y = "Count",
       caption = "Dashed red line: average oracle (p = 15)") +
  theme_minimal() +
  theme(strip.text.x = element_text(size = 11, face = "bold"),
        strip.text.y = element_text(size = 10, face = "bold"),
        panel.spacing = unit(0.5, "lines"))

print(p_hist)
ggsave("plot5_comparison_hist.png", p_hist, width = 10, height = 10, dpi = 300)
```

---

# Output 2c: Violin Plot of Selected Basis Functions

```{r violin_plot, fig.width=10, fig.height=5, eval=FALSE}
p_violin <- ggplot(plot_df, aes(x = method, y = selected)) +
  geom_hline(yintercept = 15, linetype = "dashed", color = "red", linewidth = 0.5) +
  geom_violin(aes(fill = method %in% dt_methods), trim = TRUE, scale = "width", width = 0.8) +
  geom_boxplot(width = 0.1, outlier.shape = NA, fill = "white", alpha = 0.7) +
  scale_fill_manual(values = c("TRUE" = "lightblue", "FALSE" = "gray90"), guide = "none") +
  facet_wrap(~ design, nrow = 1) +
  labs(x = NULL, y = "Selected number of basis functions") +
  theme_minimal() +
  theme(strip.text = element_text(size = 11, face = "bold"),
        axis.text.x = element_text(angle = 30, hjust = 1, size = 9))

print(p_violin)
ggsave("plot5_comparison_violin.png", p_violin, width = 10, height = 5, dpi = 300)
```

---

# Session Info

```{r session_info}
sessionInfo()
```
