---
title: "Gap Decomposition Analysis for Data Thinning"
author: "Analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 12,
  fig.height = 8,
  dpi = 300
)
```

# Overview

This analysis examines the systematic gap introduced by data thinning in model selection.

**Two key MSE quantities:**

1. **Full Oracle MSE**: Fit on full data (z), evaluate vs true θ
2. **Thinned Oracle MSE**: Fit on thinned data (z_train with mean ε×θ), evaluate vs true θ

**The Gap:**

$$\text{Gap} = \text{Thinned Oracle MSE} - \text{Full Oracle MSE}$$

This gap represents the systematic penalty from using only a fraction ε of the data for training.

**Key questions:**

- How does the gap decrease as ε increases (more data in training fold)?
- At what ε does the gap become negligible (< 5% of full oracle MSE)?
- Does the relative gap (complex vs simple models) vanish around ε ≈ 0.6-0.7?

# Data Loading

```{r load-data}
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)
library(scales)

# Custom theme
theme_set(theme_minimal(base_size = 16) +
  theme(
    panel.grid.minor = element_blank(),
    strip.background = element_rect(fill = "gray90", color = NA),
    strip.text = element_text(face = "bold", size = 18),
    legend.position = "bottom",
    axis.title = element_text(size = 16),
    axis.text = element_text(size = 14),
    legend.text = element_text(size = 14),
    legend.title = element_text(size = 15)
  ))

# Load thinned oracle MSE (S=50)
thinned_oracle <- readRDS("../results_summary/thinned_oracle_s50.RDS") %>%
  filter(n_reps_used == 1)

# Load full oracle MSE (S=50)
results <- readRDS("../results_summary/equal_allocation_results.RDS")
oracle_full <- results$oracle %>%
  filter(config %in% c("equal_30", "equal_40", "equal_50",
                       "equal_75", "equal_100", "equal_125")) %>%
  select(comp_no, config, nbasis, full_mse = mse_true)

# Verify S=50
cat("comp_no range (thinned):", range(thinned_oracle$comp_no), "\n")
cat("n_distinct comp_no:", dplyr::n_distinct(thinned_oracle$comp_no), "\n")
cat("Data loaded successfully\n")
```

# Gap Computation

```{r compute-gap}
# Join thinned and full oracle, compute gap
gap_data <- thinned_oracle %>%
  left_join(oracle_full, by = c("comp_no", "config", "nbasis")) %>%
  mutate(gap = thinned_mse - full_mse)

# Check for missing values
na_check <- gap_data %>%
  summarise(
    n_missing_gap = sum(is.na(gap)),
    n_missing_full = sum(is.na(full_mse)),
    n_missing_thinned = sum(is.na(thinned_mse))
  )

if (na_check$n_missing_gap > 0) {
  cat("WARNING: Missing gap values:", na_check$n_missing_gap, "\n")
} else {
  cat("✓ No missing values\n")
}
```

# Data Validation

```{r validation}
# Check: Thinned should be >= Full (thinning uses less data)
validation_check <- gap_data %>%
  group_by(config, epsilon) %>%
  summarise(
    n_violations = sum(gap < 0, na.rm = TRUE),
    mean_gap = mean(gap, na.rm = TRUE),
    .groups = "drop"
  )

negative_gaps <- validation_check %>%
  filter(n_violations > 0)

if (nrow(negative_gaps) > 0) {
  cat("WARNING: Some thinned MSE < full MSE (unexpected):\n")
  print(negative_gaps)
} else {
  cat("✓ All thinned MSE >= full MSE (as expected)\n")
}

# Check: Gap should decrease as epsilon increases
monotonicity_check <- gap_data %>%
  filter(nbasis == 18) %>%
  group_by(config, comp_no) %>%
  arrange(epsilon) %>%
  summarise(
    is_monotonic = all(diff(gap) < 0),
    .groups = "drop"
  )

pct_monotonic <- mean(monotonicity_check$is_monotonic) * 100

cat(sprintf("\n✓ Monotonicity (gap decreases with epsilon): %.1f%% of cases\n",
            pct_monotonic))

# Summary by config
config_summary <- gap_data %>%
  group_by(config) %>%
  summarise(
    n_rows = n(),
    mean_gap = mean(gap),
    sd_gap = sd(gap),
    mean_thinned = mean(thinned_mse),
    mean_full = mean(full_mse),
    .groups = "drop"
  )

cat("\nSummary by config (n_reps=1):\n")
print(config_summary)
```

# Figure 1: Systematic Gap vs Epsilon

Gap averaged over 50 comparisons shows how the systematic bias from thinning decreases as epsilon increases.

```{r fig1-systematic-gap, fig.width=14, fig.height=6}
# Focus on 4 nbasis values for visualization
nbasis_viz <- c(6, 18, 30, 42)

gap_summary <- gap_data %>%
  filter(nbasis %in% nbasis_viz) %>%
  group_by(config, epsilon, nbasis) %>%
  summarise(
    mean_gap = mean(gap),
    se_gap = sd(gap) / sqrt(n()),
    .groups = "drop"
  )

# Order configs by sample size and create labels
config_order <- c("equal_30", "equal_40", "equal_50", "equal_75", "equal_100", "equal_125")
config_labels <- c("n==30", "n==40", "n==50", "n==75", "n==100", "n==125")
gap_summary$config <- factor(gap_summary$config, levels = config_order, labels = config_labels)

ggplot(gap_summary, aes(x = epsilon, y = mean_gap, color = factor(nbasis))) +
  geom_line(linewidth = 1.2, alpha=0.7) +
  geom_point(size = 3) +
  facet_wrap(~ config, nrow = 2, scales = "free_y", labeller = label_parsed) +
  scale_color_brewer(palette = "Set1", name = "nbasis") +
  scale_x_continuous(breaks = seq(0.1, 0.9, 0.2)) +
  labs(
    #title = "Systematic Gap: Thinned Oracle - Full Oracle MSE (n_reps=5)",
    #subtitle = "Averaged over 20 comparisons. Fixed y-axis shows gap decreases with sample size.",
    x = "Epsilon (proportion of data in training)",
    y = "Thinning Gap",
    #caption = "Lower gap = thinning penalty smaller. Gap → 0 as ε → 1."
  ) +
  theme_minimal(base_size = 18) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(size = 20, face = "bold"),
    strip.background = element_rect(fill = "gray90", color = NA),
    axis.title = element_text(size = 18, face = "bold"),
    axis.text = element_text(size = 16),
    legend.text = element_text(size = 16),
    legend.title = element_text(size = 16, face = "bold"),
    panel.grid.minor = element_blank()
  )

ggsave("plot1_gap.png", width = 14, height = 8, dpi = 300)
```

# Figure 2: Variance of DT MSE Estimator vs Epsilon

Shows how the variance of the DT metric changes with epsilon, supporting Proposition 3.4 (variance increases for ε ≥ 0.5).

```{r fig2-dt-variance, fig.width=14, fig.height=6}
# Load DT results from paper_final (S=50)
dt_epsilon <- results$dt_1fold

dt_variance_data <- dt_epsilon %>%
  filter(
    config %in% c("equal_30", "equal_40", "equal_50",
                  "equal_75", "equal_100", "equal_125"),
    loss_function == "MSE",
    n_reps_used == 1,
    nbasis %in% c(6, 18, 30, 42)  # 4 values for visualization
  ) %>%
  group_by(config, epsilon, nbasis) %>%
  summarise(
    variance_dt = var(metric)* (n() - 1) / n(),
    .groups = "drop"
  )

# Order configs with labels (redefine in case chunk runs independently)
config_order <- c("equal_30", "equal_40", "equal_50", "equal_75", "equal_100", "equal_125")
config_labels <- c("n==30", "n==40", "n==50", "n==75", "n==100", "n==125")
dt_variance_data$config <- factor(dt_variance_data$config, levels = config_order, labels = config_labels)

# Plot: Mirror Fig 1 structure
ggplot(dt_variance_data, aes(x = epsilon, y = variance_dt,
                               color = factor(nbasis))) +
  geom_line(linewidth = 1.2, alpha=0.7) +
  geom_point(size = 3) +
  facet_wrap(~ config, nrow = 2, scales = "free_y", labeller = label_parsed) +
  scale_color_brewer(palette = "Set1", name = "nbasis") +
  scale_x_continuous(breaks = seq(0.1, 0.9, 0.2)) +
  labs(
    #title = "Variance of DT MSE Estimator vs Epsilon (n_reps=1)",
    #subtitle = "Variance computed across 20 independent comparisons. Increases for ε ≥ 0.5.",
    x = "Epsilon (proportion of data in training)",
    y = "Variance of MSE Estimator",
    #caption = "Higher variance = less stable estimates. Confirms Proposition 3.4."
  ) +
  theme_minimal(base_size = 18) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(size = 20, face = "bold"),
    strip.background = element_rect(fill = "gray90", color = NA),
    axis.title = element_text(size = 18, face = "bold"),
    axis.text = element_text(size = 16),
    legend.text = element_text(size = 16),
    legend.title = element_text(size = 16, face = "bold"),
    panel.grid.minor = element_blank()
  )

# Save figure
ggsave("plot2_variance.png", width = 14, height = 8, dpi = 300)
```

# Figure 3: The Bias-Variance Tradeoff

Combining squared thinning gap (bias²) and variance of the MSE estimator shows the fundamental tradeoff.

```{r fig3-tradeoff, fig.width=14, fig.height=8}
# Compute squared gap (bias squared) from gap_data
gap_squared <- gap_data %>%
  filter(nbasis %in% c(6, 18, 30, 42)) %>%
  group_by(config, epsilon, nbasis) %>%
  summarise(
    mean_gap_sq = mean(gap)^2,  # Square of mean gap
    .groups = "drop"
  )

# Variance data already computed
dt_variance_data <- dt_epsilon %>%
  filter(
    config %in% c("equal_30", "equal_40", "equal_50",
                  "equal_75", "equal_100", "equal_125"),
    loss_function == "MSE",
    n_reps_used == 1,
    nbasis %in% c(6, 18, 30, 42)
  ) %>%
  group_by(config, epsilon, nbasis) %>%
  summarise(
    variance_dt = var(metric) * (n() - 1) / n(),
    .groups = "drop"
  )

# Combine: Total MSE = Bias² + Variance
total_mse <- gap_squared %>%
  left_join(dt_variance_data, by = c("config", "epsilon", "nbasis")) %>%
  mutate(total_error = mean_gap_sq + variance_dt)

# Apply config labels
config_order <- c("equal_30", "equal_40", "equal_50", "equal_75", "equal_100", "equal_125")
config_labels <- c("n==30", "n==40", "n==50", "n==75", "n==100", "n==125")
total_mse$config <- factor(total_mse$config, levels = config_order, labels = config_labels)

# Find minima for each config × nbasis
minima_total <- total_mse %>%
  group_by(config, nbasis) %>%
  slice_min(total_error, n = 1)

# Plot
ggplot(total_mse, aes(x = epsilon, y = total_error, color = factor(nbasis))) +
  geom_line(linewidth = 1.2, alpha=0.7) +
  geom_point(size = 3) +
  facet_wrap(~ config, nrow = 2, scales = "free_y", labeller = label_parsed) +
  scale_color_brewer(palette = "Set1", name = "nbasis") +
  scale_x_continuous(breaks = seq(0.1, 0.9, 0.2)) +
  scale_y_continuous(labels = scales::scientific) +
  labs(
    x = "Epsilon (proportion of data in training)",
    y = expression("Thinning Gap"^2 * " + Variance")
  ) +
  theme_minimal(base_size = 18) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(size = 20, face = "bold"),
    strip.background = element_rect(fill = "gray90", color = NA),
    axis.title = element_text(size = 18, face = "bold"),
    axis.text = element_text(size = 16),
    legend.text = element_text(size = 16),
    legend.title = element_text(size = 16, face = "bold"),
    panel.grid.minor = element_blank()
  )

ggsave("plot3_tradeoff.png", width = 14, height = 8, dpi = 300)
```


```{r fig3-tradeoff-log}
# Plot
ggplot(total_mse, aes(x = epsilon, y = log(total_error), color = factor(nbasis))) +
  geom_line(linewidth = 1.2, alpha=0.7) +
  geom_point(size = 3) +
  facet_wrap(~ config, nrow = 2, scales = "free_y", labeller = label_parsed) +
  scale_color_brewer(palette = "Set1", name = "nbasis") +
  scale_x_continuous(breaks = seq(0.1, 0.9, 0.2)) +
  scale_y_continuous(labels = scales::scientific) +
  labs(
    x = "Epsilon (proportion of data in training)",
    y = expression("Thinning Gap"^2 * " + Variance")
  ) +
  theme_minimal(base_size = 18) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(size = 20, face = "bold"),
    strip.background = element_rect(fill = "gray90", color = NA),
    axis.title = element_text(size = 18, face = "bold"),
    axis.text = element_text(size = 16),
    legend.text = element_text(size = 16),
    legend.title = element_text(size = 16, face = "bold"),
    panel.grid.minor = element_blank()
  )

ggsave("plot3_tradeoff_log.png", width = 14, height = 8, dpi = 300)
```



```{r fig3-optimal-epsilon-summary}
# Summary of optimal epsilon by config and nbasis
optimal_summary <- minima_total %>%
  select(config, nbasis, epsilon_opt = epsilon, min_total_error = total_error) %>%
  arrange(config, nbasis)

cat("\nOptimal Epsilon (minimizing Bias² + Variance) by Config and Model:\n")
print(optimal_summary, n = 30)

# Overall pattern
overall_optimal <- optimal_summary %>%
  group_by(nbasis) %>%
  summarise(
    mean_eps_opt = mean(epsilon_opt),
    median_eps_opt = median(epsilon_opt),
    range = paste(min(epsilon_opt), "-", max(epsilon_opt)),
    .groups = "drop"
  )

cat("\n\nOverall Pattern Across Configs:\n")
print(overall_optimal, n = 20)
```


```{r}
overall_optimal
```


### Key Insights

1. **Interior optimum:** The bias-variance tradeoff creates an interior minimum, typically at **ε ≈ 0.4-0.6**. The U-shaped curves make this pattern visible across all configurations.

2. **Model-dependent optimal ε:**
   - Simple models (nbasis=6): Often optimal at lower ε (~0.4)
   - Moderate models (nbasis=18, 30): Typically optimal at ε ≈ 0.5 (often overlap)
   - Complex models (nbasis=42): Often optimal at higher ε (~0.5-0.6)

3. **Contrast with Section 6 recommendation (ε ≈ 0.6-0.7):**
   - The optimal ε here minimizes estimation error of MSE_full
   - Section 6's recommendation prioritizes **fair model comparison** (differential bias vanishes)
   - Different objectives → different optimal ε values

4. **Practical implication:**
   - For estimating a single model's MSE: Use ε from this analysis
   - For comparing multiple models: Use higher ε (0.6-0.7) to ensure rankings are unbiased

## Extra: Variance-Minimizing Epsilon by Model Complexity

Figure 2 shows that variance follows a U-shaped pattern with epsilon. Where is the minimum per-design?

```{r extra-variance-minima, fig.width=14, fig.height=8}
# Compute variance for each config separately
variance_by_config <- dt_epsilon %>%
  filter(
    config %in% c("equal_30", "equal_40", "equal_50",
                  "equal_75", "equal_100", "equal_125"),
    loss_function == "MSE",
    n_reps_used == 1,
    nbasis %in% c(6, 18, 30, 42)  # 4 values for visualization
  ) %>%
  group_by(config, epsilon, nbasis) %>%
  summarise(variance = var(metric) * (n() - 1) / n(), .groups = "drop")

# Find minima for each config × nbasis
minima <- variance_by_config %>%
  group_by(config, nbasis) %>%
  slice_min(variance, n = 1)

# Apply config labels
config_order <- c("equal_30", "equal_40", "equal_50", "equal_75", "equal_100", "equal_125")
config_labels <- c("n==30", "n==40", "n==50", "n==75", "n==100", "n==125")
variance_by_config$config <- factor(variance_by_config$config, levels = config_order, labels = config_labels)
minima$config <- factor(minima$config, levels = config_order, labels = config_labels)

# Plot
ggplot(variance_by_config, aes(x = epsilon, y = variance, color = factor(nbasis))) +
  geom_line(linewidth = 1.2, alpha = 0.7) +
  geom_point(size = 3) +
  geom_point(data = minima, size = 5, shape = 21, fill = "white", stroke = 1.5) +
  facet_wrap(~ config, nrow = 2, scales = "free_y", labeller = label_parsed) +
  scale_color_brewer(palette = "Set1", name = "nbasis") +
  scale_x_continuous(breaks = seq(0.1, 0.9, 0.2)) +
  scale_y_continuous(labels = scales::scientific) +
  labs(
    x = "Epsilon (proportion of data in training)",
    y = "Variance of MSE Estimator",
    caption = "Open circles mark the minimum variance for each model complexity. U-shaped pattern: low ε lacks training data, high ε lacks test data."
  ) +
  theme_minimal(base_size = 18) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(size = 20, face = "bold"),
    strip.background = element_rect(fill = "gray90", color = NA),
    axis.title = element_text(size = 18, face = "bold"),
    axis.text = element_text(size = 16),
    legend.text = element_text(size = 16),
    legend.title = element_text(size = 16, face = "bold"),
    panel.grid.minor = element_blank()
  )
```

```{r variance-minima-summary}
# Summary table of optimal epsilon by config and nbasis
optimal_summary <- minima %>%
  select(config, nbasis, epsilon_opt = epsilon, min_variance = variance) %>%
  arrange(config, nbasis)

cat("\nVariance-Minimizing Epsilon by Config and Model Complexity:\n")
print(optimal_summary, n = 30)

# Overall pattern
overall_pattern <- optimal_summary %>%
  group_by(nbasis) %>%
  summarise(
    mean_eps_opt = mean(epsilon_opt),
    range_eps = paste(min(epsilon_opt), "-", max(epsilon_opt)),
    .groups = "drop"
  )

cat("\n\nOverall Pattern (averaged across configs):\n")
print(overall_pattern, n = 20)
```

### Key Insights

1. **Consistent U-shaped pattern** across all designs and model complexities
   - Minimum typically at **ε = 0.3-0.4** regardless of sample size

2. **Simple models** (nbasis = 6, 18): Generally minimize at **ε = 0.3**
   - Less training data needed for stable estimation

3. **Complex models** (nbasis = 30, 42): Generally minimize at **ε = 0.4**
   - Need more training data to estimate many parameters

4. **The variance-optimal region (ε = 0.3-0.4) is NOT the comparison-optimal region**
   - The "Relative Gap" extra plot shows differential bias persists at low ε
   - Section 6 demonstrates ε ≈ 0.6-0.7 needed for fair comparison
   - Tradeoff: Accept ~2× variance to eliminate complexity bias

# Extra: Gap Variability Across Comparisons

Distribution of gap shows variability across the 20 independent comparisons.

```{r extra-gap-variability, fig.width=14, fig.height=6}
gap_dist <- gap_data %>%
  filter(nbasis %in% nbasis_viz) %>%
  mutate(config = factor(config, levels = config_order))

ggplot(gap_dist, aes(x = factor(epsilon), y = gap, fill = factor(nbasis))) +
  geom_boxplot(outlier.size = 0.5, alpha = 0.7) +
  facet_wrap(~ config, nrow = 2, scales = "free_y") +
  scale_fill_brewer(palette = "Set1", name = "nbasis") +
  labs(
    title = "Gap Distribution Across 20 Comparisons (n_reps=1)",
    subtitle = "Each boxplot shows spread across comparisons for fixed (config, epsilon, nbasis)",
    x = "Epsilon",
    y = "Gap (MSE units)",
    caption = "Wider boxes = more variability across comparisons"
  ) +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8)
  )
```

# Extra: Relative Gap (Complexity Penalty)

When does thinning stop differentially penalizing complex vs simple models?

```{r extra-relative-gap, fig.width=14, fig.height=5}
# Compute gap difference: complex (nbasis=42) - simple (nbasis=6)
relative_gap <- gap_data %>%
  filter(nbasis %in% c(6, 42)) %>%
  select(comp_no, config, epsilon, nbasis, gap) %>%
  pivot_wider(names_from = nbasis, values_from = gap, names_prefix = "nbasis_") %>%
  mutate(relative_gap = nbasis_42 - nbasis_6) %>%
  group_by(config, epsilon) %>%
  summarise(
    mean_relative_gap = mean(relative_gap, na.rm = TRUE),
    se_relative_gap = sd(relative_gap, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  ) %>%
  mutate(config = factor(config, levels = config_order))

ggplot(relative_gap, aes(x = epsilon, y = mean_relative_gap, color = config)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50", linewidth = 0.8) +
  geom_ribbon(
    aes(ymin = mean_relative_gap - se_relative_gap,
        ymax = mean_relative_gap + se_relative_gap, fill = config),
    alpha = 0.15, color = NA
  ) +
  scale_color_brewer(palette = "Dark2", name = "Config") +
  scale_fill_brewer(palette = "Dark2", name = "Config") +
  scale_x_continuous(breaks = seq(0.1, 0.9, 0.2)) +
  labs(
    title = "Relative Gap: Complex (nbasis=42) - Simple (nbasis=6)",
    subtitle = "When relative gap ≈ 0, thinning penalizes models equally regardless of complexity",
    x = "Epsilon",
    y = "Relative Gap (MSE units)",
    caption = "Crossing zero suggests epsilon where model ranking becomes stable under thinning"
  ) +
  theme(legend.position = "bottom")
```

# Statistical Summary Tables

## Mean Gap by Config × Epsilon

```{r table-mean-gap}
table_mean_gap <- gap_data %>%
  group_by(config, epsilon) %>%
  summarise(
    mean_gap = mean(gap),
    mean_pct_increase = mean(100 * gap / full_mse),
    .groups = "drop"
  ) %>%
  mutate(config = factor(config, levels = config_order)) %>%
  arrange(config, epsilon)

# Reshape for easier reading
table_wide <- table_mean_gap %>%
  select(config, epsilon, mean_gap) %>%
  pivot_wider(names_from = epsilon, values_from = mean_gap, names_prefix = "eps_")

table_wide
kable(table_wide, digits = 5, caption = "Mean Gap (Thinned - Full Oracle MSE) by Config and Epsilon")
```




---

# Session Info

```{r session-info}
sessionInfo()
```
