---
title: "10-Config Comprehensive Analysis: Data Thinning vs DIC/WAIC"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 6)
library(tidyverse)
library(knitr)
```

# Executive Summary

Comparison of **Data Thinning (DT)** vs **DIC/WAIC** for spatial basis function model selection across **10 sampling designs**:

- **4 Equal Allocation:** n = 40, 50, 75, 100 per PUMA
- **6 Proportional PPS:** 0.5%, 1.0%, 1.25%, 1.5%, 1.75%, 2.0% sampling fractions

**DT configs evaluated:** ε ∈ {0.3, 0.5, 0.7}, n_reps ∈ {1, 5}

---

# Data Loading

```{r load_data}
# Load aggregated 10-config data
dt_all <- readRDS("../results_multi_config/dt_all_10configs.RDS")
oracle_all <- readRDS("../results_multi_config/oracle_all_10configs.RDS")

# Get oracle selections per config/comparison
oracle_selections <- oracle_all %>%
  filter(!is.na(mse_true)) %>%
  group_by(config, comp_no) %>%
  slice_min(mse_true, n = 1, with_ties = FALSE) %>%
  select(config, comp_no, oracle_nbasis = nbasis, oracle_mse = mse_true)

cat(sprintf("Loaded %d configs with %d total comparisons\n",
            length(unique(dt_all$config)),
            nrow(oracle_selections)))

cat("\nConfigs:\n")
print(sort(unique(dt_all$config)))
```

---

# Config Characteristics

## Equivalent Sample Sizes

```{r sample_sizes}
# Load CA PUMS population
ca_pop <- readRDS("../data/ca_pums_population.rds")

# Count population by PUMA
puma_sizes <- ca_pop %>%
  group_by(PUMA) %>%
  summarise(pop = n(), .groups = "drop")

# Proportional configs
prop_configs <- tibble(
  config = c("prop_0.5pct", "prop_1pct", "prop_1p25pct", "prop_1p5pct", "prop_1p75pct", "prop_2pct"),
  samp_frac = c(0.005, 0.01, 0.0125, 0.015, 0.0175, 0.02)
)

# Sample sizes per PUMA
sample_sizes <- puma_sizes %>%
  crossing(prop_configs) %>%
  mutate(sample_n = pmax(10, round(pop * samp_frac))) %>%
  group_by(config) %>%
  summarise(
    mean_n = mean(sample_n),
    median_n = median(sample_n),
    min_n = min(sample_n),
    max_n = max(sample_n),
    .groups = "drop"
  )

# Add equal configs
equal_configs <- tibble(
  config = c("equal_40", "equal_50", "equal_75", "equal_100"),
  mean_n = c(40, 50, 75, 100),
  median_n = c(40, 50, 75, 100),
  min_n = c(40, 50, 75, 100),
  max_n = c(40, 50, 75, 100)
)

all_sample_sizes <- bind_rows(sample_sizes, equal_configs) %>%
  arrange(mean_n)

all_sample_sizes %>%
  kable(digits = 1, caption = "Equivalent Sample Sizes by Config")
```

---

# Oracle Properties by Config

## Oracle Signal Strength

```{r oracle_signal}
# Get oracle optimal nbasis per comparison
oracle_selections_signal <- oracle_all %>%
  filter(!is.na(mse_true)) %>%
  group_by(config, comp_no) %>%
  slice_min(mse_true, n = 1, with_ties = FALSE) %>%
  select(config, comp_no, oracle_nbasis = nbasis, oracle_mse = mse_true)

# Compute MSE penalties per comparison
oracle_signal <- oracle_all %>%
  filter(!is.na(mse_true)) %>%
  left_join(oracle_selections_signal, by = c("config", "comp_no")) %>%
  group_by(config, comp_no) %>%
  summarise(
    mse_min = min(mse_true),
    mse_max = max(mse_true),
    mse_range = mse_max - mse_min,
    pct_cost = (mse_max - mse_min) / mse_min * 100,
    mse_cv = sd(mse_true) / mean(mse_true) * 100,
    # Penalty for being ±6 away from optimal
    oracle_nbasis = first(oracle_nbasis),
    oracle_mse = first(oracle_mse),
    # Find MSE at oracle_nbasis ± 6 (if exists)
    mse_plus6 = mse_true[nbasis == oracle_nbasis + 6][1],
    mse_minus6 = mse_true[nbasis == oracle_nbasis - 6][1],
    .groups = "drop"
  ) %>%
  mutate(
    # Average penalty for ±6
    pct_penalty_6 = ifelse(!is.na(mse_plus6) & !is.na(mse_minus6),
                           ((mse_plus6 - oracle_mse) / oracle_mse * 100 +
                            (mse_minus6 - oracle_mse) / oracle_mse * 100) / 2,
                           ifelse(!is.na(mse_plus6),
                                  (mse_plus6 - oracle_mse) / oracle_mse * 100,
                                  ifelse(!is.na(mse_minus6),
                                         (mse_minus6 - oracle_mse) / oracle_mse * 100,
                                         NA)))
  ) %>%
  group_by(config) %>%
  summarise(
    mean_mse_range = mean(mse_range),
    mean_pct_cost = mean(pct_cost),
    mean_pct_penalty_6 = mean(pct_penalty_6, na.rm = TRUE),
    mean_mse_cv = mean(mse_cv),
    .groups = "drop"
  ) %>%
  left_join(all_sample_sizes %>% select(config, mean_n), by = "config") %>%
  arrange(mean_n) %>%
  select(-mean_n)

oracle_signal %>%
  kable(digits = c(0, 5, 2, 2, 2),
        caption = "Oracle Signal Strength: mean_pct_cost = % increase worst vs best; mean_pct_penalty_6 = avg % penalty at ±6 from optimal")
```

## Oracle MSE Curves

```{r oracle_curves, fig.height=10}
oracle_all %>%
  filter(!is.na(mse_true)) %>%
  ggplot(aes(x = nbasis, y = mse_true)) +
  geom_line(aes(group = comp_no), alpha = 0.3, color = "darkgreen") +
  geom_point(data = oracle_selections,
             aes(x = oracle_nbasis, y = oracle_mse),
             color = "red", size = 1) +
  facet_wrap(~config, ncol = 3, scales = "free_y") +
  labs(title = "Oracle MSE Curves by Config",
       subtitle = "Red points = optimal nbasis for each comparison",
       x = "nbasis",
       y = "MSE") +
  theme_minimal()
```

---

# Method Performance: DT vs DIC/WAIC

## Computing MAD (Mean Absolute Deviation)

```{r compute_mad}
# DIC selections
dic_selections <- oracle_all %>%
  filter(!is.na(DIC)) %>%
  group_by(config, comp_no) %>%
  slice_min(DIC, n = 1, with_ties = FALSE) %>%
  select(config, comp_no, dic_nbasis = nbasis)

# WAIC selections
waic_selections <- oracle_all %>%
  filter(!is.na(WAIC)) %>%
  group_by(config, comp_no) %>%
  slice_min(WAIC, n = 1, with_ties = FALSE) %>%
  select(config, comp_no, waic_nbasis = nbasis)

# DT selections (filter to MSE loss only, select best epsilon/n_reps per config)
dt_selections <- dt_all %>%
  filter(method_name == "dt_1fold", loss_function == "MSE") %>%
  group_by(config, comp_no, epsilon, n_reps_used) %>%
  slice_min(metric, n = 1, with_ties = FALSE) %>%
  select(config, comp_no, epsilon, n_reps_used, dt_nbasis = nbasis)

# Compute MAD for DIC
dic_mad <- dic_selections %>%
  left_join(oracle_selections, by = c("config", "comp_no")) %>%
  mutate(deviation = abs(dic_nbasis - oracle_nbasis)) %>%
  group_by(config) %>%
  summarise(MAD = mean(deviation), method = "DIC", .groups = "drop")

# Compute MAD for WAIC
waic_mad <- waic_selections %>%
  left_join(oracle_selections, by = c("config", "comp_no")) %>%
  mutate(deviation = abs(waic_nbasis - oracle_nbasis)) %>%
  group_by(config) %>%
  summarise(MAD = mean(deviation), method = "WAIC", .groups = "drop")

# Compute MAD for all DT configs (filter to n=1 and n=5 only)
dt_mad <- dt_selections %>%
  filter(n_reps_used %in% c(1, 5)) %>%
  left_join(oracle_selections, by = c("config", "comp_no")) %>%
  mutate(deviation = abs(dt_nbasis - oracle_nbasis)) %>%
  group_by(config, epsilon, n_reps_used) %>%
  summarise(MAD = mean(deviation), .groups = "drop") %>%
  mutate(method = sprintf("DT ε=%.1f n=%d", epsilon, n_reps_used))
```

```{r combine_mad}
# Combine all MADs (for downstream use)
all_mad <- bind_rows(
  dic_mad,
  waic_mad,
  dt_mad %>% select(config, MAD, method)
)
```

---

# Oracle MSE Penalty Analysis

## Computing Oracle MSE Penalties

```{r oracle_penalties}
# Get Oracle MSE at DIC's selected nbasis
dic_oracle_penalty <- dic_selections %>%
  left_join(oracle_all %>% select(config, comp_no, nbasis, mse_true),
            by = c("config", "comp_no", "dic_nbasis" = "nbasis")) %>%
  rename(dic_mse = mse_true) %>%
  left_join(oracle_selections, by = c("config", "comp_no")) %>%
  mutate(
    pct_penalty = (dic_mse - oracle_mse) / oracle_mse * 100
  ) %>%
  group_by(config) %>%
  summarise(
    mean_pct_penalty = mean(pct_penalty),
    method = "DIC",
    .groups = "drop"
  )

# Get Oracle MSE at WAIC's selected nbasis
waic_oracle_penalty <- waic_selections %>%
  left_join(oracle_all %>% select(config, comp_no, nbasis, mse_true),
            by = c("config", "comp_no", "waic_nbasis" = "nbasis")) %>%
  rename(waic_mse = mse_true) %>%
  left_join(oracle_selections, by = c("config", "comp_no")) %>%
  mutate(
    pct_penalty = (waic_mse - oracle_mse) / oracle_mse * 100
  ) %>%
  group_by(config) %>%
  summarise(
    mean_pct_penalty = mean(pct_penalty),
    method = "WAIC",
    .groups = "drop"
  )

# Get Oracle MSE at DT's selected nbasis (filtered to n=1, 5)
dt_oracle_penalty <- dt_selections %>%
  filter(n_reps_used %in% c(1, 5)) %>%
  left_join(oracle_all %>% select(config, comp_no, nbasis, mse_true),
            by = c("config", "comp_no", "dt_nbasis" = "nbasis")) %>%
  rename(dt_mse = mse_true) %>%
  left_join(oracle_selections, by = c("config", "comp_no")) %>%
  mutate(
    pct_penalty = (dt_mse - oracle_mse) / oracle_mse * 100
  ) %>%
  group_by(config, epsilon, n_reps_used) %>%
  summarise(
    mean_pct_penalty = mean(pct_penalty),
    .groups = "drop"
  ) %>%
  mutate(method = sprintf("DT ε=%.1f n=%d", epsilon, n_reps_used))

# Combine all penalties
all_oracle_penalties <- bind_rows(
  dic_oracle_penalty,
  waic_oracle_penalty,
  dt_oracle_penalty %>% select(config, mean_pct_penalty, method)
)
```

## Best Methods: MAD vs Oracle Penalty

```{r best_methods_comparison}
# Best by MAD
best_by_mad <- all_mad %>%
  group_by(config) %>%
  slice_min(MAD, n = 1, with_ties = FALSE) %>%
  select(config, winner_mad = method, best_mad = MAD)

# Best by Oracle Penalty
best_by_oracle_penalty <- all_oracle_penalties %>%
  group_by(config) %>%
  slice_min(mean_pct_penalty, n = 1, with_ties = FALSE) %>%
  select(config, winner_penalty = method, best_penalty = mean_pct_penalty)

# Combine side-by-side (ordered by sample size)
best_methods_combined <- best_by_mad %>%
  left_join(best_by_oracle_penalty, by = "config") %>%
  left_join(all_sample_sizes %>% select(config, mean_n), by = "config") %>%
  arrange(mean_n) %>%
  select(-mean_n)

best_methods_combined %>%
  kable(digits = 2, caption = "Winner by MAD vs Oracle MSE Penalty")
```

---

# DT vs DIC Gap Analysis

## Visualization: DT vs DIC MAD Gap

```{r gap_plot_mad, fig.height=7}
# Find best DT MAD per config
best_dt_mad <- dt_mad %>%
  group_by(config) %>%
  slice_min(MAD, n = 1, with_ties = FALSE) %>%
  select(config, best_dt_method = method, best_dt_mad = MAD)

# Join with DIC MAD
gap_analysis_mad <- dic_mad %>%
  rename(dic_mad = MAD) %>%
  select(config, dic_mad) %>%
  left_join(best_dt_mad, by = "config") %>%
  mutate(
    gap = best_dt_mad - dic_mad,
    winner = ifelse(gap < 0, "DT", "DIC")
  )

gap_analysis_mad %>%
  mutate(config = fct_reorder(config, gap)) %>%
  ggplot(aes(x = config, y = gap, fill = winner)) +
  geom_col() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  coord_flip() +
  scale_fill_manual(values = c("DT" = "steelblue", "DIC" = "coral")) +
  labs(title = "Best DT vs DIC: MAD Gap",
       subtitle = "Negative = DT wins, Positive = DIC wins",
       x = "Config",
       y = "MAD Gap (DT - DIC)",
       fill = "Winner") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Visualization: DT vs DIC Oracle Penalty Gap

```{r gap_plot_oracle, fig.height=7}
# Find best DT Oracle penalty per config
best_dt_penalty <- dt_oracle_penalty %>%
  group_by(config) %>%
  slice_min(mean_pct_penalty, n = 1, with_ties = FALSE) %>%
  select(config, best_dt_method = method, best_dt_penalty = mean_pct_penalty)

# Join with DIC Oracle penalty
gap_analysis_oracle <- dic_oracle_penalty %>%
  rename(dic_penalty = mean_pct_penalty) %>%
  select(config, dic_penalty) %>%
  left_join(best_dt_penalty, by = "config") %>%
  mutate(
    gap = best_dt_penalty - dic_penalty,
    winner = ifelse(gap < 0, "DT", "DIC")
  )

gap_analysis_oracle %>%
  mutate(config = fct_reorder(config, gap)) %>%
  ggplot(aes(x = config, y = gap, fill = winner)) +
  geom_col() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  coord_flip() +
  scale_fill_manual(values = c("DT" = "steelblue", "DIC" = "coral")) +
  labs(title = "Best DT vs DIC: Oracle MSE Penalty Gap",
       subtitle = "Negative = DT wins, Positive = DIC wins",
       x = "Config",
       y = "Oracle Penalty Gap % (DT - DIC)",
       fill = "Winner") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

---

# Epsilon Sensitivity Analysis

## Metric 1: Mean Absolute Deviation (MAD)

MAD measures the average distance (in nbasis units) between the selected model and the oracle optimal model.

```{r epsilon_curves_mad, fig.height=10, fig.width=12}
# Prepare MAD data across epsilon values
dt_mad_by_epsilon <- dt_selections %>%
  left_join(oracle_selections, by = c("config", "comp_no")) %>%
  mutate(deviation = abs(dt_nbasis - oracle_nbasis)) %>%
  group_by(config, epsilon, n_reps_used) %>%
  summarise(MAD = mean(deviation), .groups = "drop") %>%
  mutate(method = "DT")

# DIC/WAIC MAD (already computed earlier)
dic_mad_for_plot <- dic_mad %>%
  mutate(epsilon = NA, n_reps_used = NA)

waic_mad_for_plot <- waic_mad %>%
  mutate(epsilon = NA, n_reps_used = NA)

# Combine
all_mad_plot <- bind_rows(
  dt_mad_by_epsilon,
  dic_mad_for_plot,
  waic_mad_for_plot
)

# Order configs by sample size
config_levels <- all_sample_sizes %>%
  arrange(mean_n) %>%
  pull(config)

# Plot: DT n=5 only with DIC/WAIC references
all_mad_plot %>%
  filter(method == "DT" & n_reps_used == 5 | method %in% c("DIC", "WAIC")) %>%
  mutate(config = factor(config, levels = config_levels)) %>%
  ggplot(aes(x = epsilon, y = MAD)) +
  geom_hline(data = all_mad_plot %>% filter(method == "DIC") %>%
               mutate(config = factor(config, levels = config_levels)),
             aes(yintercept = MAD, color = "DIC"),
             linetype = "dashed", linewidth = 0.8) +
  geom_hline(data = all_mad_plot %>% filter(method == "WAIC") %>%
               mutate(config = factor(config, levels = config_levels)),
             aes(yintercept = MAD, color = "WAIC"),
             linetype = "dashed", linewidth = 0.8) +
  geom_line(data = all_mad_plot %>% filter(method == "DT", n_reps_used == 5) %>%
              mutate(config = factor(config, levels = config_levels)),
            aes(color = "DT n=5"), linewidth = 1.2) +
  geom_point(data = all_mad_plot %>% filter(method == "DT", n_reps_used == 5) %>%
               mutate(config = factor(config, levels = config_levels)),
             aes(color = "DT n=5"), size = 2.5) +
  facet_wrap(~config, ncol = 3, scales = "free_y") +
  scale_color_manual(
    values = c("DT n=5" = "steelblue", "DIC" = "coral", "WAIC" = "darkgoldenrod2"),
    breaks = c("DT n=5", "DIC", "WAIC")
  ) +
  labs(
    title = "Method Performance: Mean Absolute Deviation from Oracle Optimal",
    subtitle = "DT with n=5 averaging (solid line) vs DIC/WAIC (dashed lines). Lower is better. Ordered by sample size.",
    x = "Epsilon (ε)",
    y = "MAD (nbasis units)",
    color = "Method"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )
```

## Metric 2: Oracle MSE Penalty

Oracle MSE Penalty measures the percentage increase in MSE from selecting a suboptimal model compared to the oracle optimal model.

```{r epsilon_curves_penalty, fig.height=10, fig.width=12}
# Prepare data for oracle penalty plot
dt_oracle_penalty_full <- dt_selections %>%
  left_join(oracle_all %>% select(config, comp_no, nbasis, mse_true),
            by = c("config", "comp_no", "dt_nbasis" = "nbasis")) %>%
  rename(dt_mse = mse_true) %>%
  left_join(oracle_selections, by = c("config", "comp_no")) %>%
  mutate(pct_increase = (dt_mse - oracle_mse) / oracle_mse * 100) %>%
  group_by(config, epsilon, n_reps_used) %>%
  summarise(
    mean_pct_increase = mean(pct_increase),
    se_pct_increase = sd(pct_increase) / sqrt(n()),
    .groups = "drop"
  ) %>%
  mutate(method = "DT")

# DIC penalties
dic_oracle_penalty_full <- dic_oracle_penalty %>%
  select(config, mean_pct_increase = mean_pct_penalty) %>%
  mutate(method = "DIC", epsilon = NA, n_reps_used = NA)

# WAIC penalties
waic_oracle_penalty_full <- waic_oracle_penalty %>%
  select(config, mean_pct_increase = mean_pct_penalty) %>%
  mutate(method = "WAIC", epsilon = NA, n_reps_used = NA)

# Combine all
all_penalties_plot <- bind_rows(
  dt_oracle_penalty_full,
  dic_oracle_penalty_full,
  waic_oracle_penalty_full
)

# Plot: DT n=5 only with DIC/WAIC references
all_penalties_plot %>%
  filter(method == "DT" & n_reps_used == 5 | method %in% c("DIC", "WAIC")) %>%
  mutate(config = factor(config, levels = config_levels)) %>%
  ggplot(aes(x = epsilon, y = mean_pct_increase)) +
  geom_hline(data = all_penalties_plot %>% filter(method == "DIC") %>%
               mutate(config = factor(config, levels = config_levels)),
             aes(yintercept = mean_pct_increase, color = "DIC"),
             linetype = "dashed", linewidth = 0.8) +
  geom_hline(data = all_penalties_plot %>% filter(method == "WAIC") %>%
               mutate(config = factor(config, levels = config_levels)),
             aes(yintercept = mean_pct_increase, color = "WAIC"),
             linetype = "dashed", linewidth = 0.8) +
  geom_line(data = all_penalties_plot %>% filter(method == "DT", n_reps_used == 5) %>%
              mutate(config = factor(config, levels = config_levels)),
            aes(color = "DT n=5"), linewidth = 1.2) +
  geom_point(data = all_penalties_plot %>% filter(method == "DT", n_reps_used == 5) %>%
               mutate(config = factor(config, levels = config_levels)),
             aes(color = "DT n=5"), size = 2.5) +
  facet_wrap(~config, ncol = 3, scales = "free_y") +
  scale_color_manual(
    values = c("DT n=5" = "steelblue", "DIC" = "coral", "WAIC" = "darkgoldenrod2"),
    breaks = c("DT n=5", "DIC", "WAIC")
  ) +
  labs(
    title = "Method Performance: Percentage Increase in MSE from Oracle Model",
    subtitle = "DT with n=5 averaging (solid line) vs DIC/WAIC (dashed lines). Lower is better. Ordered by sample size.",
    x = "Epsilon (ε)",
    y = "Percentage Increase in MSE (%)",
    color = "Method"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )
```

## Best Epsilon Per Config

```{r best_epsilon}
best_epsilon_table <- all_penalties_plot %>%
  filter(method == "DT", n_reps_used == 5) %>%
  group_by(config) %>%
  slice_min(mean_pct_increase, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  left_join(all_sample_sizes %>% select(config, mean_n), by = "config") %>%
  arrange(mean_n) %>%
  select(config, best_epsilon = epsilon, pct_increase = mean_pct_increase) %>%
  left_join(
    dic_oracle_penalty_full %>% select(config, DIC_pct = mean_pct_increase),
    by = "config"
  ) %>%
  mutate(
    dt_vs_dic = pct_increase - DIC_pct,
    winner = ifelse(dt_vs_dic < 0, "DT", "DIC")
  )

best_epsilon_table %>%
  kable(digits = 2,
        caption = "Best epsilon (n=5) per config with comparison to DIC. Negative dt_vs_dic = DT wins.")
```

## Mean Performance Across Configs

```{r epsilon_summary}
epsilon_summary <- all_penalties_plot %>%
  filter((method == "DT" & n_reps_used == 5) | method == "DIC") %>%
  group_by(method, epsilon) %>%
  summarise(
    mean_pct = mean(mean_pct_increase),
    sd_pct = sd(mean_pct_increase),
    .groups = "drop"
  ) %>%
  arrange(mean_pct)

epsilon_summary %>%
  kable(digits = 2,
        caption = "Mean percentage increase in MSE across 10 configs")
```

---

# Competitive Landscape

## Method Rankings: Detailed

```{r method_rankings, results='asis'}
# Order configs by sample size
config_order <- all_sample_sizes %>%
  arrange(mean_n) %>%
  pull(config)

for (cfg in config_order) {
  cat(sprintf("\n### %s\n\n", cfg))

  cat("**By MAD:**\n\n")
  rankings_mad <- all_mad %>%
    filter(config == cfg) %>%
    arrange(MAD) %>%
    mutate(rank = 1:n()) %>%
    select(rank, method, MAD)

  print(kable(rankings_mad, digits = 2))
  cat("\n\n")

  cat("**By Oracle MSE Penalty:**\n\n")
  rankings_penalty <- all_oracle_penalties %>%
    filter(config == cfg) %>%
    arrange(mean_pct_penalty) %>%
    mutate(rank = 1:n()) %>%
    select(rank, method, penalty = mean_pct_penalty)

  print(kable(rankings_penalty, digits = 2))
  cat("\n\n")
}
```


---

# Session Info

```{r session}
sessionInfo()
```
