---
title: "Epsilon Sensitivity Analysis for Spatial Basis Fixed Effects Model"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 6)
library(tidyverse)
library(knitr)
```


```{r load_data}
# Load aggregated 8-config data
dt_all <- readRDS("../results_multi_config/dt_epsilon_8configs.RDS")
oracle_all <- readRDS("../results_multi_config/oracle_epsilon_8configs.RDS")

# FOCUS: n_reps = 5 only (most stable with maximum averaging)
dt_all <- dt_all %>% filter(n_reps_used == 5)

# Order config by sample size
config_levels <- c(
  "equal_20", "equal_30", "equal_40", "equal_50", "equal_75",
  "equal_100", "equal_125", "equal_150"
)
dt_all <- dt_all %>% mutate(config = factor(config, levels = config_levels))
oracle_all <- oracle_all %>% mutate(config = factor(config, levels = config_levels))

# Get oracle selections per config/comparison
oracle_selections <- oracle_all %>%
  filter(!is.na(mse_true)) %>%
  group_by(config, comp_no) %>%
  slice_min(mse_true, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(config, comp_no, oracle_nbasis = nbasis, oracle_mse = mse_true)

# cat(sprintf("Loaded %d configs (n_reps=5 only)\n", length(unique(dt_all$config))))
# cat(sprintf("Total observations: %d\n", nrow(dt_all)))
# cat(sprintf(
#   "Sample sizes: %s\n",
#   paste(sort(unique(as.numeric(str_extract(dt_all$config, "[0-9]+")))), collapse = ", ")
# ))
# cat(sprintf(
#   "Epsilon grid: %s\n",
#   paste(sort(unique(dt_all$epsilon)), collapse = ", ")
# ))
```

```{r compute_mad}
# Get DT selections (minimum metric value)
dt_selections <- dt_all %>%
  group_by(config, comp_no, epsilon, loss_function) %>%
  slice_min(metric, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(config, comp_no, epsilon, loss_function, selected_nbasis = nbasis)

# Join with oracle selections
dt_deviations <- dt_selections %>%
  left_join(oracle_selections, by = c("config", "comp_no")) %>%
  mutate(
    deviation = selected_nbasis - oracle_nbasis,
    abs_deviation = abs(deviation),
    sample_size = as.numeric(str_extract(config, "[0-9]+"))
  )

# Compute MAD per config × epsilon × loss
mad_summary <- dt_deviations %>%
  group_by(config, sample_size, epsilon, loss_function) %>%
  summarise(
    MAD = mean(abs_deviation),
    mean_dev = mean(deviation),
    sd_dev = sd(deviation),
    .groups = "drop"
  )

# Find optimal epsilon per config × loss
optimal_eps <- mad_summary %>%
  group_by(config, sample_size, loss_function) %>%
  slice_min(MAD, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(config, sample_size, loss_function, optimal_eps = epsilon, optimal_MAD = MAD)

#cat(sprintf("Computed MAD for %d epsilon × config × loss combinations\n", nrow(mad_summary)))
```

# Overview 

This analysis investigates how **optimal epsilon for data thinning** varies with equal allocation sample sizes (n=20 to n=150) using a Fay-Herriot model that uses fixed-effect spatial basis functions. 

I tried **equal allocation** sampling by PUMA. I varied this from n=20 to 150. 
All results shown for **n_reps=5** (maximum averaging, most stable).

### Oracle Signal Doesn't Always Improve with sample size 

We can see the observed data oracle MSE curves below.  We can see that the oracle signal doesn't actually improve with sample size. It gets 'noisier' after a certain sweet spot around 40-75. 


```{r oracle_curves, fig.height=8}
oracle_all %>%
  filter(!is.na(mse_true)) %>%
  ggplot(aes(x = nbasis, y = mse_true)) +
  geom_line(aes(group = comp_no), alpha = 0.3, color = "darkgreen") +
  geom_point(
    data = oracle_selections,
    aes(x = oracle_nbasis, y = oracle_mse),
    color = "red", size = 1
  ) +
  facet_wrap(~config, ncol = 3, scales = "free_y") +
  labs(
    title = "Oracle MSE Curves by Config",
    subtitle = "Red points = optimal nbasis for each comparison",
    x = "nbasis",
    y = "MSE"
  ) +
  theme_minimal()
```


The Optimal Epsilon actually follows this general trajectory of the oracle signal:

```{r epsilon_trajectory, fig.width=10, fig.height=5}
optimal_eps %>% 
ggplot(aes(x = sample_size, y = optimal_eps, color = loss_function, group = loss_function)) +
  geom_line(alpha=0.8) +
  geom_point(alpha=0.7) +
  scale_color_brewer(palette = "Set1", name = "Loss Function") +
  scale_y_continuous(breaks = seq(0.1, 0.9, 0.1), limits = c(0, 1)) +
  labs(
    title = "Optimal Epsilon vs Sample Size",
    x = "Sample Size (n per PUMA)",
    y = "Optimal Epsilon"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 12),
    legend.position = "bottom"
  )
```

Optimal epsilon follows a non-monotonic pattern driven by oracle signal quality:

- **n=20:** eps ≈ 0.1 (too little data, unreliable, probably not enough for CLT)
- **n=30-40:** eps ≈ 0.7-0.9 (peak - best oracle signal)
- **n=50-100:** eps ≈ 0.6-0.7 (plateau - stable range)
- **n=125-150:** eps ≈ 0.4-0.5 (decline - flatter oracle curves)

**The Story:** As sample size increases, direct estimates improve but oracle curves flatten. This requires adjusting epsilon: larger samples paradoxically need MORE validation data (lower epsilon) to detect differences between models.



### Epsilon vs MAD of Basis Selected (compared to Oracle)

```{r mad_curves, fig.width=8, fig.height=6}
# Annotate optimal epsilon
mad_with_optimal <- mad_summary %>%
  left_join(optimal_eps %>% select(sample_size, loss_function, optimal_eps),
    by = c("sample_size", "loss_function")
  ) %>% 
  filter(loss_function=="MSE")

mad_with_optimal %>% 
  ggplot(aes(x = epsilon, y = MAD, group=factor(sample_size))) +
  geom_line(size=1, color='blue') +
  geom_point(size=1, color='blue') +
  geom_point(
    data = mad_with_optimal %>% filter(abs(epsilon - optimal_eps) < 0.01),
    size = 3, shape = 21, stroke = 2, 
  ) +
  facet_wrap(~factor(sample_size), scales='free_y') +
  labs(
    x = "Epsilon",
    y = "MAD (Mean Absolute Deviation from Oracle)"
  ) +
  #theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    strip.text = element_text(face = "bold", size = 14),
    legend.position = "right"
  )
```

**Key observations:**

- **Small samples (n=20):** Very steep curves on left (big overfitting penalty)
- **Mid-range (n=30-50):** Big underfitting penalty
- **Large samples (n=100-125):** More U-Shaped 



# Best Average Performance

Which epsilon performs best **on average** across all sample sizes?

```{r average_performance}
# Average MAD across all sample sizes for each epsilon
avg_mad_by_epsilon <- dt_deviations %>%
  group_by(epsilon) %>%
  summarise(
    avg_MAD = mean(abs_deviation),
    sd_MAD = sd(abs_deviation),
    median_MAD = median(abs_deviation),
    n_configs = n_distinct(config),
    .groups = "drop"
  ) %>%
  arrange(avg_MAD)


kable(avg_mad_by_epsilon,
  digits = 2,
  col.names = c("Epsilon", "Avg MAD", "SD", "Median MAD", "N Configs")
)

best_eps <- avg_mad_by_epsilon$epsilon[1]
best_mad <- avg_mad_by_epsilon$avg_MAD[1]
second_eps <- avg_mad_by_epsilon$epsilon[2]
second_mad <- avg_mad_by_epsilon$avg_MAD[2]
safe_range <- avg_mad_by_epsilon %>%
  filter(avg_MAD <= best_mad + 0.2)
```

**Interpretation:**

- **epsilon = `r best_eps`** has the best average performance across all sample sizes
- **epsilon = 0.4-0.7** all perform relatively similiarly 
- Avoid extremes: eps ≤ 0.2 or eps ≥ 0.9 perform substantially worse



### Epsilon Range by Sample Size

```{r epsilon_range}
epsilon_range <- mad_summary %>%
  group_by(sample_size) %>%
  summarise(
    min_MAD = min(MAD),
    max_MAD = max(MAD),
    range_MAD = max_MAD - min_MAD,
    best_eps = epsilon[which.min(MAD)],
    worst_eps = epsilon[which.max(MAD)],
    .groups = "drop"
  )
kable(epsilon_range,
  digits = 2,
  col.names = c("Sample Size", "Min MAD", "Max MAD", "Range", "Best ε", "Worst ε")
)
```

Range of MAD shows how much epsilon choice matters. Larger ranges = more sensitive to epsilon selection.

### Concluding Thoughts


Just like the discussion in this paper [this paper](https://arxiv.org/abs/2511.12698), I think that the optimal epsilon depends on 

1. model complexity 

2. the signal in the data itself. 

In this analysis, we're only able to vary 2. We need to look at 1 as well. 
