---
title: "Results from Equal Allocation n=75 Design"
output: html_document
date: "2025-11-26"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
suppressPackageStartupMessages({
  library(tidyverse)
})
```

```{r load_result, warning=FALSE}
# Load results
setwd("~/dev/dt_design_based_comp")
results <- readRDS("_results_ca_full_comparison_equal75/results.RDS") %>%
  mutate(epsilon=round(epsilon, 1)) %>%
  filter((metric_type !="predictive_NLL") | is.na(metric_type)) %>%
 # filter(model!="D.Est") %>%
 filter(!is.na(nbasis))

# Get oracle selections per comparison
oracle_selections <- results %>%
  filter(method_name == "oracle") %>%
  group_by(comp_no) %>%
  slice_min(metric, n = 1, with_ties = FALSE) %>%
  select(comp_no, oracle_nbasis = nbasis)

# Join with all methods and calculate deviations
deviations <- results %>%
  filter(method_name !="oracle") %>%
  group_by(method, comp_no, method_name, metric_type, epsilon, n_reps_used) %>%
  slice_min(metric, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(method, comp_no, method_name, metric_type, epsilon, n_reps_used, selected_nbasis = nbasis) %>%
  left_join(oracle_selections, by = "comp_no") %>%
  mutate(deviation = selected_nbasis - oracle_nbasis,
         abs_deviation = abs(deviation))
```

# Background

We are benchmarking against the Observed Data Oracle MSE (the realized MSE for a given dataset).


**OD Oracle Optimal Basis Distribution (C=50 Comparions)**

```{r fig.height=3, fig.width=5}
table(oracle_selections$oracle_nbasis) %>% barplot(main="")
```
More concentrated compared to proportional PPS sampling.

Here are the oracle MSE curves across 50 direct estimates (the minimum is shown as a point):


```{r fig.height=3, fig.width=5}
tmp_df = results %>%
  filter(method_name == "oracle")

tmp_df %>%
  ggplot(aes(x=nbasis, y=metric)) +
  geom_line(aes(group=comp_no), alpha=0.4, color="darkgreen") +
  geom_point(data=tmp_df %>% group_by(comp_no) %>% filter(min(metric)==metric),
             color="darkgreen") #+
  # stat_summary(aes(group = 1), fun = mean, geom = "line",
  #              color = "blue", linewidth = 1.2)
```


***

# ‼ Overall Results ‼

DT single-fold Restricted to 5 repeats

```{r}
# Summary statistics by method (collapsing epsilon/n_reps for DT)
overall <- deviations %>%
  filter((n_reps_used==5) |  is.na(epsilon)) %>%
  group_by(method) %>%
  summarise(
    mean_dev = round(mean(deviation), 1),
    sd_dev = round(sd(deviation), 1),
    mean_abs_dev = round(mean(abs_deviation), 1),
    median_abs_dev = round(median(abs_deviation), 1),
    .groups = "drop"
  ) %>%
  arrange(mean_abs_dev)
overall %>% 
  knitr::kable()
```


We just look at epsilon=0.6 and 0.7

```{r}
# Summary statistics by method (collapsing epsilon/n_reps for DT)
overall <- deviations %>%
  filter((epsilon %in% c(0.6, 0.7)) |  is.na(epsilon)) %>%
  group_by(method) %>%
  summarise(
    mean_dev = round(mean(deviation), 1),
    sd_dev = round(sd(deviation), 1),
    mean_abs_dev = round(mean(abs_deviation), 1),
    median_abs_dev = round(median(abs_deviation), 1),
    .groups = "drop"
  ) %>%
  arrange(mean_abs_dev)
overall %>% 
  knitr::kable()
```


```{r fig.height=8, fig.width=9}
# Prepare data for each method
# 1. Oracle (OD-Oracle MSE)
oracle <- results %>%
  filter(method_name == "oracle") %>%
  select(comp_no, nbasis, value = metric) %>%
  mutate(method = "OD-Oracle MSE")

# 2. WAIC
waic <- results %>%
  filter(method_name == "waic") %>%
  select(comp_no, nbasis, value = metric) %>%
  mutate(method = "WAIC")

# 3. DIC
dic <- results %>%
  filter(method_name == "dic") %>%
  select(comp_no, nbasis, value = metric) %>%
  mutate(method = "DIC")

# 4. DT 5-fold (use MSE loss)
dt5 <- results %>%
  filter(method_name == "dt_5fold", metric_type == "MSE") %>%
  select(comp_no, nbasis, value = metric) %>%
  mutate(method = "DT 5-fold")

# dt5_nll <- results %>%
#   filter(method_name == "dt_5fold", metric_type == "plugin_NLL") %>%
#   select(comp_no, nbasis, value = metric) %>%
#   mutate(method = "DT 5-fold")

# 5. DT 1-fold (eps=0.3, 3 reps, MSE loss)
dt1_02 <- results %>%
  filter(method_name == "dt_1fold", epsilon == 0.7, n_reps_used == 5, metric_type == "MSE") %>%
  select(comp_no, nbasis, value = metric) %>%
  mutate(method = "DT 1-fold (eps=0.7)")

# 6. DT 1-fold (eps=0.5, 5 reps, NLL loss)
dt1_04 <- results %>%
  filter(method_name == "dt_1fold", epsilon == 0.5, n_reps_used == 5, metric_type == "MSE") %>%
  select(comp_no, nbasis, value = metric) %>%
  mutate(method = "DT 1-fold (eps=0.5)")

# 7. DT 1-fold (eps=0.5, 5 reps, NLL loss)
dt1_06 <- results %>%
  filter(method_name == "dt_1fold", epsilon == 0.5, n_reps_used == 5, metric_type == "plugin_NLL") %>%
  select(comp_no, nbasis, value = metric) %>%
  mutate(method = "DT 1-fold - NLL (eps=0.5)")

# 8. ESIM Standard
esim_std <- results %>%
  filter(method_name == "esim_standard") %>%
  select(comp_no, nbasis, value = metric) %>%
  mutate(method = "ESIM Standard")

# 9. ESIM Data Fission
esim_fis <- results %>%
  filter(method_name == "esim_fission") %>%
  select(comp_no, nbasis, value = metric) %>%
  mutate(method = "ESIM Data Fission")

# Combine all methods
all_methods <- bind_rows(oracle, waic, dic, dt5,
                         dt1_02, dt1_04, dt1_06, esim_std, esim_fis)

# Get number of comparisons
n_comp <- length(unique(all_methods$comp_no))

# Create faceted plot by METHOD (each method in its own panel)
# Individual comparisons = light gray lines, Average = bold blue line
ggplot(all_methods, aes(x = nbasis, y = value)) +
  geom_line(aes(group = comp_no), alpha = 0.25, color = "gray60", linewidth = 0.5) +
  stat_summary(aes(group = 1), fun = mean, geom = "line",
               color = "blue", linewidth = 1.2) +
  facet_wrap(~method, ncol = 3, scales = "free_y") +
  labs(
    title = sprintf("All Methods: Individual vs Average (%d comparisons)", n_comp),
    subtitle = "Gray = individual comparisons, Blue = average. Raw values (free y-axis per method)",
    x = "Number of basis functions",
    y = "Score"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 12),
    strip.text = element_text(face = "bold", size = 10),
    axis.text = element_text(size = 8)
  )

```

### Thoughts on results

DIC/WAIC perform well with equal allocation. DT shows better performance with higher $\epsilon$ values (0.5-0.7 range) in this design.


## DT Single-Fold: Impact of Epsilon & Repeats


```{r }
# Also show best DT config by epsilon/reps
dt_single_all <- deviations %>%
  filter(method_name %in% c("dt_1fold")) %>%
  group_by(method_name, metric_type, epsilon, n_reps_used) %>%
  summarise(
    mean_dev= round(mean(deviation), 1),
    sd_dev=round(sd(deviation), 1),
    mean_abs_dev = round(mean(abs_deviation), 1),
    median_abs_dev = round(median(abs_deviation), 1),
    .groups = "drop"
  ) %>%
  arrange(mean_abs_dev)
```



```{r fig.width=9, fig.height=3}
dt_single_all %>%
  mutate(n_reps_used=as.factor(n_reps_used)) %>%
  ggplot(aes(x=epsilon, y=mean_abs_dev, colour = n_reps_used)) + geom_point() + geom_line() +
  geom_point(size=4, color="navy", shape=5, data=dt_single_all %>%
               group_by(method_name, metric_type,  n_reps_used) %>%
               filter(min(mean_abs_dev)==mean_abs_dev)) +
  facet_wrap(metric_type~.)  +
  ylab("Mean Abs. Deviation from OD-Oracle Basis")+
  theme_bw()+
  ggtitle("DT Single-Fold MAD of Basis Selection Across Epsilon & Repeats")
```


**Key patterns:**
- Optimal $\epsilon$ around 0.6-0.7 for MSE loss
- Higher $\epsilon$ helps across both loss functions
- More reps generally improve performance  

Below is a plot showing the distribution of the basis deviations from the oracle for DT with MSE. DT underestimates the number of basis functions needed, especially for lower epsilon values where less information is available for model fitting.

```{r key_boxplot}
deviations %>%
  filter(method_name =="dt_1fold" & metric_type=="MSE") %>%
  mutate(epsilon=as.factor(epsilon)) %>%
  ggplot(aes(x=epsilon, y=deviation)) +
  geom_boxplot() +
  geom_abline(slope = 0, linetype=2, color="red") +
  facet_grid(n_reps_used~.) +
  ggtitle("Deviations from OD-Oracle Basis across 50 Comparisons (faceted by repeats)")
```


## Looking at deviations from other methods

```{r fig.width=8, fig.height=3}
deviations %>%
  filter(is.na(epsilon)) %>%
  ggplot(aes(x=method, y=deviation)) +
  geom_boxplot() +
  geom_abline(slope = 0, linetype=2, color="red") +
  ggtitle("Deviations from OD-Oracle Basis across 50 Comparisons")
```

DIC/WAIC show near-zero mean deviation with moderate variance. ESIM methods and DT 5-fold show varying patterns of bias and variance.
