---
title: "Section 2: Spatial Smoothing Illustration"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 14, fig.height = 5)
library(tidyverse)
library(sf)
library(tigris)
library(latex2exp)
options(tigris_use_cache = TRUE)
```

# Configuration

```{r config}
# Sampling design to visualize
CONFIG <- "equal_75"  # Options: equal_40, equal_50, equal_75, equal_100, prop_0.5pct, etc.

# Which comparison to use
COMP_NO <- 13
# 10, 2, 12
# Basis function ranks to compare
NBASIS_VALUES <- c(6, 42)
```

# Load Data

```{r load-data}
# Map config names to results directories
config_dirs <- list(
  equal_30 = "_results_equal30",
  equal_40 = "_results_equal40",
  equal_50 = "_results_equal50",
  equal_75 = "_results_equal75",
  equal_100 = "_results_equal100",
  equal_125 = "_results_equal125",
  prop_0.75pct = "_results_prop0.75pct",
  prop_1p25pct = "_results_prop1p25pct",
  prop_1p75pct = "_results_prop1p75pct"
)

results_dir <- file.path("..", config_dirs[[CONFIG]])
chains_file <- file.path(results_dir, sprintf("comparison_%03d/fit_on_z/chains.RDS", COMP_NO))

# Load pre-fitted models
chains <- readRDS(chains_file)

# Extract spatial smooth (theta - w) for selected nbasis values
models <- lapply(NBASIS_VALUES, function(nb) {
  chain_idx <- which(sapply(chains, function(x) x$nbasis) == nb)

  if (length(chain_idx) == 0) {
    stop(sprintf("nbasis = %d not found in chains", nb))
  }

  theta_mcmc <- chains[[chain_idx]]$chain$theta
  w_mcmc <- chains[[chain_idx]]$chain$w
  puma_ids <- chains[[chain_idx]]$puma_ids

  # Compute X*beta = theta - w (spatial smooth without IID noise)
  spatial_smooth <- colMeans(theta_mcmc) - colMeans(w_mcmc)

  list(
    nbasis = nb,
    puma_ids = puma_ids,
    spatial_smooth = spatial_smooth
  )
})

# Get CA PUMA boundaries from Census
ca_pumas <- pumas(state = "CA", cb = TRUE, year = 2020)

# Create estimates dataframe
estimates_df <- bind_rows(
  lapply(models, function(m) {
    tibble(
      PUMA = m$puma_ids,
      spatial_smooth = m$spatial_smooth,
      nbasis = m$nbasis
    )
  })
)

# Join with shapefile
map_data <- ca_pumas %>%
  mutate(PUMA = PUMACE20) %>%
  inner_join(estimates_df, by = "PUMA") %>%
  mutate(nbasis = factor(nbasis, levels = NBASIS_VALUES))

cat(sprintf("Loaded %d PUMA boundaries with estimates for nbasis: %s\n",
            nrow(map_data) / length(NBASIS_VALUES),
            paste(NBASIS_VALUES, collapse = ", ")))
```

# Figure 0: Spatial Smoothing Across Model Complexity

```{r figure0-spatial-map, fig.width=12, fig.height=8}
# LA County approximate bounding box
la_bbox <- c(xmin = -118.7, xmax = -117.8, ymin = 33.6, ymax = 34.2)

# Get CA bounding box for positioning inset
ca_bbox <- st_bbox(map_data)

# Create separate plots for each nbasis with inset
plots_with_inset <- lapply(unique(map_data$nbasis), function(nb) {
  # Filter data for this nbasis
  data_nb <- map_data %>% filter(nbasis == nb)

  # Main plot
  p_main <- ggplot(data_nb) +
    geom_sf(aes(fill = spatial_smooth), color = "gray30", linewidth = 0.1) +
    # Add rectangle showing zoom area
    geom_rect(aes(xmin = la_bbox["xmin"], xmax = la_bbox["xmax"],
                  ymin = la_bbox["ymin"] + 0.1, ymax = la_bbox["ymax"]),
              fill = NA, color = "red", linewidth = 0.8, inherit.aes = FALSE) +
    scale_fill_gradient(low = "white", high = "darkblue",
                       limits = range(map_data$spatial_smooth),
                       name = TeX("$X\\beta$")) +
    # scale_fill_viridis_c(
    #   option = "viridis",
    #   limits = range(map_data$spatial_smooth),
    #   name = TeX("$X\\beta$")
    # ) +
    labs(title = paste0("p = ", nb)) +
    theme_minimal(base_size = 12) +
    theme(
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      panel.grid = element_blank(),
      plot.title = element_text(size = 24, face = "bold"),
      legend.position = "right",
      legend.text = element_text(size = 14),
      legend.title = element_text(size = 16)
    )

  # Zoom inset (smaller)
  p_inset <- ggplot(data_nb) +
    geom_sf(aes(fill = spatial_smooth), color = "gray30", linewidth = 0.3) +
    coord_sf(xlim = c(la_bbox["xmin"], la_bbox["xmax"]),
             ylim = c(la_bbox["ymin"], la_bbox["ymax"]),
             expand = FALSE) +
    scale_fill_gradient(low = "white", high = "darkblue",
                       limits = range(map_data$spatial_smooth)) +
    # scale_fill_viridis_c(
    #   option = "viridis",
    #   limits = range(map_data$spatial_smooth)
    # ) +
    theme_void() +
    theme(
      panel.background = element_rect(fill = "white", color = NA),
      legend.position = "none",
      plot.background = element_rect(fill = "white", color = NA)
    )

  # Convert inset to grob
  inset_grob <- ggplotGrob(p_inset)

  # Inset position coordinates
  inset_xmin <- ca_bbox["xmin"] + 0.6 * (ca_bbox["xmax"] - ca_bbox["xmin"])
  inset_xmax <- ca_bbox["xmax"]
  inset_ymin <- ca_bbox["ymin"] + 0.4 * (ca_bbox["ymax"] - ca_bbox["ymin"])
  inset_ymax <- ca_bbox["ymin"] + 0.67 * (ca_bbox["ymax"] - ca_bbox["ymin"])

  # Add inset to main plot, then draw red border on top (avoids clipping)
  p_with_inset <- p_main +
    annotation_custom(
      grob = inset_grob,
      xmin = inset_xmin, xmax = inset_xmax,
      ymin = inset_ymin, ymax = inset_ymax
    ) +
    annotate("rect", xmin = inset_xmin, xmax = inset_xmax,
             ymin = inset_ymin, ymax = inset_ymax,
             fill = NA, color = "red", linewidth = 0.8)

  p_with_inset
})

# Combine plots side by side using patchwork
library(patchwork)
combined <- wrap_plots(plots_with_inset, nrow = 1) +
  plot_annotation(
    #title = "Spatial Smoothing with Los Angeles Inset (red box)",
    #theme = theme(plot.title = element_text(size = 24, face = "bold"))
  )
combined
```

**Key observation:** The LA inset reveals the spatial smoothing effect. With nbasis = 6, the entire LA region is highly homogeneous (strong spatial pooling). With nbasis = 60, neighborhood-level variation emerges - but note it's still much smoother than raw data since we're showing X*beta without IID noise.


```{r}
ggsave("plot0_la_zoom.png", combined, width = 12, height = 8, dpi = 300)
```


# Figure 0b: SF Bay Area Zoom

```{r figure0b-sf-zoom, fig.width=14, fig.height=6}
# SF Bay Area approximate bounding box
sf_bbox <- c(xmin = -122.6, xmax = -121.7, ymin = 37.35, ymax = 38.05)

# Get CA bounding box for positioning inset
ca_bbox <- st_bbox(map_data)

# Create separate plots for each nbasis with inset
plots_with_inset_sf <- lapply(unique(map_data$nbasis), function(nb) {
  # Filter data for this nbasis
  data_nb <- map_data %>% filter(nbasis == nb)

  # Main plot
  p_main <- ggplot(data_nb) +
    geom_sf(aes(fill = spatial_smooth), color = "gray30", linewidth = 0.1) +
    # Add rectangle showing zoom area
    geom_rect(aes(xmin = sf_bbox["xmin"], xmax = sf_bbox["xmax"],
                  ymin = sf_bbox["ymin"], ymax = sf_bbox["ymax"]),
              fill = NA, color = "red", linewidth = 0.8, inherit.aes = FALSE) +
    scale_fill_gradient(low = "white", high = "darkblue",
                       limits = range(map_data$spatial_smooth)) +
    # scale_fill_viridis_c(
    #   option = "viridis",
    #   limits = range(map_data$spatial_smooth),
    #   name = TeX("$X\\beta$")
    # ) +
    labs(title = paste0("p = ", nb)) +
    theme_minimal(base_size = 12) +
    theme(
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      panel.grid = element_blank(),
      plot.title = element_text(size = 16, face = "bold"),
      legend.position = "right"
    )

  # Zoom inset (smaller)
  p_inset <- ggplot(data_nb) +
    geom_sf(aes(fill = spatial_smooth), color = "gray30", linewidth = 0.3) +
    coord_sf(xlim = c(sf_bbox["xmin"], sf_bbox["xmax"]),
             ylim = c(sf_bbox["ymin"], sf_bbox["ymax"]),
             expand = FALSE) +
    scale_fill_gradient(low = "white", high = "darkblue",
                       limits = range(map_data$spatial_smooth)) +
    # scale_fill_viridis_c(
    #   option = "viridis",
    #   limits = range(map_data$spatial_smooth)
    # ) +
    theme_void() +
    theme(
      panel.background = element_rect(fill = "white", color = NA),
      legend.position = "none",
      plot.background = element_rect(fill = "white", color = NA)
    )

  # Convert inset to grob
  inset_grob <- ggplotGrob(p_inset)

  # Inset position coordinates
  inset_xmin <- ca_bbox["xmin"] + 0.6 * (ca_bbox["xmax"] - ca_bbox["xmin"])
  inset_xmax <- ca_bbox["xmax"]
  inset_ymin <- ca_bbox["ymin"] + 0.62 * (ca_bbox["ymax"] - ca_bbox["ymin"])
  inset_ymax <- ca_bbox["ymin"] + 0.92 * (ca_bbox["ymax"] - ca_bbox["ymin"])

  # Add inset to main plot, then draw red border on top (avoids clipping)
  p_with_inset <- p_main +
    annotation_custom(
      grob = inset_grob,
      xmin = inset_xmin, xmax = inset_xmax,
      ymin = inset_ymin, ymax = inset_ymax
    ) +
    annotate("rect", xmin = inset_xmin, xmax = inset_xmax,
             ymin = inset_ymin, ymax = inset_ymax,
             fill = NA, color = "red", linewidth = 0.8)

  p_with_inset
})

# Combine plots side by side using patchwork
combined_sf <- wrap_plots(plots_with_inset_sf, nrow = 1) +
  plot_annotation(
    #title = "Spatial Smoothing with SF Bay Area Inset (red box)",
    theme = theme(plot.title = element_text(size = 16, face = "bold"))
  )

combined_sf

#ggsave("plot0_sf_zoom.png", combined_sf, width = 14, height = 6, dpi = 300)
```

**Key observation:** The SF Bay Area inset demonstrates spatial smoothing across the dense urban core. With nbasis = 6, SF/Oakland/San Jose appear as one smooth region. With nbasis = 60, distinct patterns emerge for each city and neighborhood.

**Interpretation:** Maps show **X*beta = theta - w** (spatial smooth without IID noise), isolating the spatial basis contribution.

Model selection involves choosing the number of spatial basis functions (nbasis), which controls spatial smoothing:

- **Low complexity (nbasis = 6)**: Strong spatial smoothing - neighboring areas have very similar values
- **High complexity (nbasis = 60)**: Weak smoothing - more local variation between neighbors

By removing the IID random effects (w), these maps clearly show how the spatial basis functions pool information across space. The optimal choice balances bias (underfitting with too few basis functions) and variance (overfitting with too many).

# Summary Statistics

```{r summary-stats}
summary_stats <- bind_rows(
  lapply(models, function(m) {
    tibble(
      nbasis = m$nbasis,
      mean = mean(m$spatial_smooth),
      sd = sd(m$spatial_smooth),
      range = max(m$spatial_smooth) - min(m$spatial_smooth)
    )
  })
)

summary_stats
```

As nbasis increases: SD ↑, Range ↑ (less smoothing, more local variation)

---

# Session Info

```{r session-info}
sessionInfo()
```
