---
title: "Gibbs Sampler Diagonstics"
author: "Sho Kawano"
date: "`r Sys.Date()`"
output:   
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
  # To run the validation:
  # source("test_gibbs_vs_stan.R")

  # To load pre-computed results (if available):

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 10, fig.height = 6
)
library(tidyverse)
library(ggplot2)
library(gridExtra)
library(bayesplot)

set.seed(12345)
bayesplot_theme_set(theme_minimal())
```

# Overview

This document demonstrates one chain of the **Spatial Basis Function Fay-Herriot Gibbs Sampler** using North Carolina county test data.

**Mean Model:**
$$\theta_i = X_i'\beta + v_i + w_i$$

- $v = S\eta$, $\eta \sim N(0, \tau^2 I_r)$ (spatial random effects via basis functions)
- $w \sim N(0, \sigma^2 I_m)$ (IID random effects)
- $z_i | \theta_i \sim N(\theta_i, d_i)$ (observed direct estimates with known variances)

# Data Setup

## Load Test Data

```{r load-data}
# Load NC test data (100 counties)
load("nc_testdata.RDA")

cat(sprintf(
  "Test data loaded:
  Areas: %d (NC counties)
  Adjacency matrix: %d × %d
  Avg neighbors per county: %.2f

Available variables:
  %s\n",
  nrow(all_data),
  nrow(A), ncol(A),
  mean(rowSums(A)),
  paste(names(all_data), collapse = ", ")
))
```

## Prepare Model Inputs

```{r prepare-inputs}
# Response: log poverty
z <- log(all_data$povPerc)
names(z) <- all_data$fips

# Variances: from delta method
d <- all_data$povPercSE^2 / all_data$povPerc^2
names(d) <- all_data$fips

# Create predictor matrix
# Select relevant demographic/economic covariates
X <- all_data %>%
  select(
    income, degree, white, black, hispanic,
    assistance, female_only, no_car, age.60_plus
  ) %>%
  mutate(intercept = 1) %>%
  select(intercept, everything()) %>%
  as.matrix()

rownames(X) <- all_data$fips

# Ensure alignment
stopifnot(all(rownames(X) == names(z)))
stopifnot(all(rownames(X) == rownames(A)))

cat(sprintf(
  "Response variable (poverty percentage):
  Mean: %.2f%%
  SD: %.2f%%
  Median SE: %.2f%%
  Range: %.2f%% to %.2f%%

Predictors created:
  Dimensions: %d × %d
  Variables: %s

✓ All data aligned correctly\n",
  mean(z), sd(z), sqrt(median(d)), min(z), max(z),
  nrow(X), ncol(X),
  paste(colnames(X), collapse = ", ")
))
```

# Run Gibbs Sampler

## Model Configuration

```{r model-config}
# MCMC settings
ndesired <- 2000
nburn <- 5000
nthin <- 1

# Number of spatial basis functions to test
nbasis_values <- seq(from = 10, to = 35, by = 5)

cat(sprintf(
  "MCMC Configuration:
  Burn-in: %d iterations
  Desired samples: %d iterations
  Thinning: %d
  Total iterations: %d

Testing %d basis function configurations:
  %s\n",
  nburn, ndesired, nthin, nburn + ndesired * nthin,
  length(nbasis_values),
  paste(nbasis_values, collapse = ", ")
))
```

## Run Models

```{r run-models, warning=TRUE, results='hide'}
source("spatial_basis_fh.R")

# Store results
results <- list()
runtimes <- numeric(length(nbasis_values))

for (i in seq_along(nbasis_values)) {
  nb <- nbasis_values[i]

  cat("\n=== Running model with nbasis =", nb, "===\n")

  start_time <- Sys.time()

  fit <- spatial_basis_fh(
    X = X,
    y = z,
    d.var = d,
    nbasis = nb,
    A = A,
    ndesired = ndesired,
    nburn = nburn,
    nthin = nthin,
    hyp = list(a = 1, b = 1, c = 1, d = 1),
    verbose = TRUE
  )

  end_time <- Sys.time()
  runtime <- as.numeric(difftime(end_time, start_time, units = "secs"))

  results[[paste0("nbasis_", nb)]] <- fit
  runtimes[i] <- runtime

  cat("Runtime:", round(runtime, 2), "seconds\n")
}
```

```{r runtime-summary}
runtime_df <- data.frame(
  nbasis = nbasis_values,
  runtime_sec = round(runtimes, 2)
)
colnames(runtime_df) <- c("Basis Functions", "Runtime (sec)")

cat("Model Runtimes:\n\n")
print(runtime_df, row.names = FALSE)
```

## Gibbs Sampler Validation vs. Stan

For comprehensive validation of the Gibbs sampler against Stan, see **`models/test_gibbs_vs_stan.R`**.

That script compares:
- Posterior means and SDs for σ², τ², and regression coefficients
- Small area means (θ) across all areas
- Runtime comparisons across multiple basis function configurations
- Convergence diagnostics (Rhat, ESS, divergences)

```{r load-validation-results}
# To run the validation:
# source("models/test_gibbs_vs_stan.R")

# To load pre-computed results (if available):
if (file.exists("../results/gibbs_vs_stan_timing.csv")) {
  validation_results <- read.csv("../results/gibbs_vs_stan_timing.csv")

  timing <- validation_results %>%
    filter(success == TRUE) %>%
    select(nbasis, method, time_sec, mean_tau_sq, mean_sigma_sq)

  if (nrow(timing) > 0) {
    # Create wide format table for timing comparison
    timing_wide <- timing %>%
      pivot_wider(
        id_cols = nbasis,
        names_from = method,
        values_from = c(time_sec, mean_tau_sq, mean_sigma_sq),
        names_glue = "{method}_{.value}"
      ) %>%
      mutate(
        speedup = `R Gibbs_time_sec` / `Stan_time_sec`
      ) %>%
      select(
        nbasis,
        `R Gibbs_time_sec`,
        `Stan_time_sec`,
        speedup,
        `R Gibbs_mean_tau_sq`,
        `Stan_mean_tau_sq`,
        `R Gibbs_mean_sigma_sq`,
        `Stan_mean_sigma_sq`
      ) %>%
      arrange(nbasis)

    # Format column names for display
    timing_display <- timing_wide
    colnames(timing_display) <- c(
      "nbasis",
      "Gibbs (s)",
      "Stan (s)",
      "Speedup",
      "Gibbs τ²",
      "Stan τ²",
      "Gibbs σ²",
      "Stan σ²"
    )

    cat("Gibbs vs. Stan Timing and Estimates:\n\n")
    print(timing_display, digits = 3, row.names = FALSE)
  }
} else {
  cat("Run models/test_gibbs_vs_stan.R to generate validation results.\n")
}
```

# Diagnostics

## Trace Plots (small area means - thetas)

```{r trace-plots, fig.height=8}
# Create trace plots using bayesplot for selected models
selected_nbasis <- c(15, 25, 35)

for (nb in selected_nbasis) {
  fit <- results[[paste0("nbasis_", nb)]]

  # Combine variance components and first few betas
  mcmc_draws <- cbind(
    theta = fit$theta[, c(3, 40, 90, 99)]
  )

  p <- mcmc_trace(mcmc_draws) +
    ggtitle(paste0("MCMC Trace - nbasis = ", nb))
  print(p)
}
```


## Trace Plots (parameters)

```{r trace-plots2, fig.height=8}
for (nb in selected_nbasis) {
  fit <- results[[paste0("nbasis_", nb)]]

  # Combine variance components and first few betas
  mcmc_draws <- cbind(
    sigma.sq = fit$sigma.sq[, 1],
    tau.sq = fit$tau.sq[, 1],
    fit$beta[, 1:min(4, ncol(fit$beta))]
  )

  p <- mcmc_trace(mcmc_draws) +
    ggtitle(paste0("MCMC Trace - nbasis = ", nb))
  print(p)
}
```

## Posterior Densities

```{r posterior-densities, fig.height=5}
# Create density plots using bayesplot for selected models
for (nb in selected_nbasis) {
  fit <- results[[paste0("nbasis_", nb)]]

  # Variance components and betas
  mcmc_draws <- cbind(
    sigma.sq = fit$sigma.sq[, 1],
    tau.sq = fit$tau.sq[, 1],
    fit$beta[, 1:min(4, ncol(fit$beta))]
  )

  p <- mcmc_dens(mcmc_draws) +
    ggtitle(paste0("Posterior Densities - nbasis = ", nb))
  print(p)
}
```



## Autocorrelation Plots

```{r autocorrelation, fig.height=6}
# Check autocorrelation for variance components
for (nb in selected_nbasis) {
  fit <- results[[paste0("nbasis_", nb)]]

  mcmc_draws <- cbind(
    theta = fit$theta[, c(7, 39)], # choose random area
    sigma.sq = fit$sigma.sq[, 1],
    tau.sq = fit$tau.sq[, 1]
  )

  p <- mcmc_acf(mcmc_draws, lags = 20) +
    ggtitle(paste0("Autocorrelation - nbasis = ", nb))
  print(p)
}
```



## Posterior Means Comparison

```{r posterior-means-comparison, fig.height=3, fig.width=9}
# Extract posterior means of theta for each model
theta_posteriors <- lapply(seq_along(nbasis_values), function(i) {
  nb <- nbasis_values[i]
  fit <- results[[paste0("nbasis_", nb)]]
  colMeans(fit$theta)
})

# Create comparison data frame
theta_df <- data.frame(
  area = rownames(X),
  direct = z,
  se = sqrt(d)
)

for (i in seq_along(nbasis_values)) {
  theta_df[[paste0("nbasis_", nbasis_values[i])]] <- theta_posteriors[[i]]
}

# Plot: Direct SE vs. Residual (z - posterior mean)
plot_list <- list()
for (nb in selected_nbasis) {
  col_name <- paste0("nbasis_", nb)

  # Calculate residual: direct estimate minus posterior mean
  plot_data <- theta_df %>%
    mutate(residual = direct - .data[[col_name]])

  p <- ggplot(plot_data, aes(x = se, y = residual)) +
    geom_point(alpha = 0.5, color = "steelblue") +
    geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
    labs(
      title = paste0("nbasis = ", nb),
      x = "D.Est SE",
      y = "D.Est - Posterior Mean"
    ) +
    theme_minimal()

  plot_list[[length(plot_list) + 1]] <- p
}

do.call(grid.arrange, c(plot_list, ncol = 3))
```

## Shrinkage Analysis

```{r shrinkage, fig.height=3, fig.width=8}
# Calculate shrinkage for each model
shrinkage_df <- data.frame(
  area = rownames(X),
  direct = z,
  se = sqrt(d)
)

for (i in seq_along(nbasis_values)) {
  nb <- nbasis_values[i]
  col_name <- paste0("nbasis_", nb)
  shrinkage_df[[col_name]] <- theta_posteriors[[i]]
  shrinkage_df[[paste0("shrinkage_", nb)]] <- (theta_posteriors[[i]] - z) / z * 100
}

# Shrinkage vs. SE (show selected models)
plot_list <- list()
for (nb in selected_nbasis) {
  shrink_col <- paste0("shrinkage_", nb)

  p <- ggplot(shrinkage_df, aes_string(x = "se", y = shrink_col)) +
    geom_point(alpha = 0.5, color = "darkgreen") +
    labs(
      title = paste0("nbasis = ", nb),
      x = "Direct Estimate SE (%)",
      y = "Shrinkage (% Difference from D.Est)"
    ) +
    theme_minimal()

  plot_list[[length(plot_list) + 1]] <- p
}

do.call(grid.arrange, c(plot_list, ncol = 3))
```

# Summary Statistics

```{r summary-stats, echo=FALSE}
# Comparison table
comparison_df <- data.frame(
  nbasis = nbasis_values,
  runtime_sec = runtimes
)

# Add variance component estimates
for (i in seq_along(nbasis_values)) {
  nb <- nbasis_values[i]
  fit <- results[[paste0("nbasis_", nb)]]
  comparison_df$sigma_sq_mean[i] <- mean(fit$sigma.sq[, 1])
  comparison_df$tau_sq_mean[i] <- mean(fit$tau.sq[, 1])
  comparison_df$var_ratio[i] <- mean(fit$tau.sq[, 1]) / mean(fit$sigma.sq[, 1])
}

# Add mean absolute shrinkage
for (i in seq_along(nbasis_values)) {
  nb <- nbasis_values[i]
  shrink_col <- paste0("shrinkage_", nb)
  comparison_df$mean_shrinkage_pct[i] <- mean(shrinkage_df[[shrink_col]])
}

# Format and display
comparison_formatted <- comparison_df
comparison_formatted[, 2:6] <- round(comparison_formatted[, 2:6], 3)
colnames(comparison_formatted) <- c(
  "Basis Functions", "Runtime (s)",
  "σ² (mean)", "τ² (mean)", "τ²/σ²",
  "Mean Shrinkage (% Difference from D.Est)"
)

print(comparison_formatted, row.names = FALSE)
```

## Variance Components vs. Number of Basis Functions

```{r variance-vs-nbasis, fig.height=6}
# Create plot showing how variance components change with nbasis
var_plot_df <- comparison_df %>%
  select(nbasis, sigma_sq_mean, tau_sq_mean) %>%
  pivot_longer(
    cols = c(sigma_sq_mean, tau_sq_mean),
    names_to = "component",
    values_to = "variance"
  ) %>%
  mutate(component = ifelse(component == "sigma_sq_mean",
    "σ² (IID)",
    "τ² (Spatial)"
  ))

p_var <- ggplot(var_plot_df, aes(x = nbasis, y = variance, color = component)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(
    title = "Variance Components vs. Number of Basis Functions",
    x = "Number of Basis Functions",
    y = "Posterior Mean Variance",
    color = "Component"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

print(p_var)
```

# Random Effects Visualization

## Load Spatial Geometry

```{r load-geometry}
library(sf)

# Load NC county geometry
nc_geom <- readRDS("nc_testdata_w_geom.RDS")

# Create reference table for joining
ref_table <- data.frame(
  area = 1:nrow(all_data),
  fips = all_data$fips
)
```

## Spatial Random Effects (v)

```{r spatial-random-effects, fig.height=8}
# Extract posterior means of spatial random effects for selected models
v_df <- data.frame(fips = all_data$fips)

for (nb in selected_nbasis) {
  fit <- results[[paste0("nbasis_", nb)]]
  v_df[[paste0("nbasis_", nb)]] <- colMeans(fit$v)
}

# Join with geometry
v_spatial <- nc_geom %>%
  inner_join(v_df, by = c("GEOID" = "fips"))

# Convert to long format for faceting
v_long <- v_spatial %>%
  pivot_longer(
    cols = starts_with("nbasis_"),
    names_to = "model",
    values_to = "v_effect",
    names_prefix = "nbasis_"
  ) %>%
  mutate(model = paste0("nbasis = ", model))

# Create spatial plot
p_v <- ggplot(v_long) +
  geom_sf(aes(fill = v_effect)) +
  scale_fill_distiller(palette = "RdBu", direction = 1) +
  facet_wrap(~model, ncol = 1) +
  labs(
    title = "Spatial Random Effects (v = S·η)",
    fill = "Effect"
  ) +
  theme_void() +
  theme(
    legend.position = "right",
    strip.text = element_text(size = 10)
  )

print(p_v)
```


